---
title: "Ordination of all samples"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    toc_depth: 4
    df_print: paged
author: "Sabrina Pankey"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE,results="hide")
```


```{r load, echo=FALSE, include=FALSE} 
mainDir<-"~/Box Sync/MPL/Lesser DOB sponges/coevolution_analyses/pipeline4/"
setwd(mainDir)
outDir<-"02_tree_results/"
dir.exists(file.path(mainDir, outDir))
library(phyloseq)
library(RColorBrewer) 
library(colorspace)
library(dplyr)
library(ggplot2)
library(ape)
load("01_results/psR.rds")
meta16<-data.frame(sample_data(psR)) %>% select(DNAid,voucher,Species_consensus,species)
  
load(paste0(outDir,"treeitems.Rds")) #loads, tr and colsC==Order colors --tree is filtered DOBsamples tree,no refs
  keep<-sample_names(psR)[(sample_data(psR)$DNAid %in% tr$tip.label)]
  psR<-prune_samples(keep,psR)

load("01_results/ps4.rds") #loads ps4--filtered samples, not rarefied, for subsamples for each clade
  keep<-sample_names(ps4)[(sample_data(ps4)$DNAid %in% tr$tip.label)]
  ps4<-prune_samples(keep,ps4)
```

##Data subsets: 

 - sample tree, 
 - species time-tree, 
 - species character tree
 
```{r summarycollection,results="markdown"}
#number samples x site
#numer orders/species x site

load(paste0(outDir,"spongespeciesTimeTree.OGless.Rds")) #loads timetr,cladecols,cladelist,treemeta.u,colsO,colsF
samp<-data.frame(sample_data(ps4))
samp<-subset(samp, select=c("Species_consensus","locale","phyloOrder","site"))
  samp<-samp[!is.na(samp$Species_consensus),]

#ggtree tip order
  timetreetips<-ggtree::get_taxa_name(ggtree::ggtree(timetr))
  
    #ape tip order
  is_tip <- timetr$edge[,2] <= length(timetr$tip.label)
    ord <- timetr$edge[is_tip, 2]; timetreetipsA<-timetr$tip.label[ord]
samp$Species_consensus<-factor(samp$Species_consensus,levels=timetreetips[timetreetips %in% samp$Species_consensus])
samp$phyloOrder<-factor(samp$phyloOrder,levels=names(colsC)[names(colsC) %in% samp$phyloOrder])
spbars<-unlist(lapply(samp$phyloOrder,FUN=function(x) colsC[names(colsC)==x]))

spbars1<-cbind(samp,spbars) %>% group_by(Species_consensus,spbars) %>% summarize()
spbars2<-as.character(spbars1$spbars)

#plot1: summarize to species, facet by locale, color by order, order x axis by phylo
p1<-ggplot(samp,aes(x=Species_consensus,fill=phyloOrder)) + geom_bar() + scale_fill_manual(values=spbars) +theme(legend.position="none",axis.text.x = element_text(angle = 90,size=4,hjust=1))
p1 + facet_wrap(~locale,scales="free")

p2<-ggplot(samp,aes(x=phyloOrder,fill=Species_consensus)) + geom_bar(position = "stack", color = "black") + scale_fill_manual(values=spbars2) +theme(legend.position="none",axis.text.x = element_text(angle = 90,size=4,hjust=1))
p2+ facet_wrap(~locale,scales="free")
p3<-ggplot(samp,aes(x=locale,fill=phyloOrder)) + geom_bar() + scale_fill_manual(values=colsC) +theme(legend.position="none")
p3
p4<-ggplot(samp,aes(x=locale,fill=Species_consensus)) + geom_bar(position = "stack", color = "black") + scale_fill_manual(values=spbars2) + theme(legend.position="none") 
p4

library(colorspace)
hma.col<-rainbow_hcl(12)[2]; lma.col<-rainbow_hcl(12)[7]

#summarize 3 tree datasets
barcodedn<-samp %>% group_by(Species_consensus) %>% summarize(n=n()) %>% filter(!is.na(Species_consensus))
#read in character data
chardat<-read.csv("02_input_spongetree/chars_by_species.csv")
types<-chardat %>% filter(!is.na(consensusmod))
chem<-chardat %>% filter(!is.na(LohmeanPellets))

`%notin%` <- Negate(`%in%`)


pdf(paste0(outDir,"summary_tree.pdf"),height=30, width=8)
plot(timetr,show.tip.label=F,no.margin=T,x.lim=1000)


for (i in types$Species_consensus){
  if (i %in% timetr$tip.label){
    t<-types$consensusmod[which(types$Species_consensus==i)]
    if(t=="HMA"){
  tiplabels(text=t,tip=which(timetr$tip.label==i),col=hma.col,cex=0.4,frame="none",offset=10)}
     if(t=="LMA"){
  tiplabels(text=t,tip=which(timetr$tip.label==i),col=lma.col,cex=0.4,frame="none",offset=10)}
  }}
for (i in chem$Species_consensus){
  if (i %in% timetr$tip.label){
  tiplabels(pch=5,tip=which(timetr$tip.label==i),cex=0.4,frame="none",offset=25)}
}

tiplabels(tip=which(timetr$tip.label %notin% barcodedn$Species_consensus),text=timetr$tip.label[which(timetr$tip.label %notin% barcodedn$Species_consensus)],col="gray70",cex=0.4,frame="none",adj=c(0,0.5),offset = 45)
tiplabels(tip=which(timetr$tip.label %in% barcodedn$Species_consensus),timetr$tip.label[which(timetr$tip.label %in% barcodedn$Species_consensus)],col="black",cex=0.4,frame="none",adj=c(0,0.5),offset = 45)
for (i in barcodedn$Species_consensus){
  if (i %in% timetr$tip.label){
    n<-barcodedn$n[which(barcodedn$Species_consensus==i)]
  tiplabels(text=n,tip=which(timetr$tip.label==i),col="red",cex=0.4,frame="none",offset=35)
  }}
dev.off()
sampleN<-nrow(samp)
dobspp<-length(barcodedn$Species_consensus)
sptips<-Ntip(timetr)
reftipsN<-length(which(treemeta.u$Species_consensus %notin% samp$Species_consensus))
nrow(chem)
oc<-length(which(chem$Species_consensus %in% barcodedn$Species_consensus))
nrow(types)
ot<-length(which(types$Species_consensus %in% barcodedn$Species_consensus))
```

Datasets analysed: 
samples barcoded host+micro: `r sampleN`
barcoded samples placed into `r dobspp` species
Number of species in species tree: `r sptips`
Number of species tip from GenBank:`r reftipsN`
Number of species with published HMA/LMA states: `r nrow(types)`
Number of species with published HMA/LMA states, overlapping with DOB `r ot `.

Number of species with published chemical defense data `r nrow(chem)`.
Number of species with published chemical defense data, overlapping with DOB `r oc`.

Save sample summary table
- samples by collection site, species
```{r saveSampTable}
samp<-data.frame(sample_data(ps4))

test<-samp  %>% group_by(Species_consensus,locale) %>% summarize(n=n()) 
test
#unique(samp$Species_consensus)
#samp[is.na(samp$Species_consensus),]
tab<-reshape2::dcast(test, Species_consensus~locale)
tab<-tab[!is.na(tab$Species_consensus),]
tab<-tab[!is.na(tab$Species_consensus),]
tab[(is.na(tab))]<-0
sum(colSums(tab[,2:5]))
tab2<-left_join(tab,treemeta.u) %>% select(Species_consensus,BZ,CU,KY,SX,WPDOrder,phyloOrder)

write.csv(tab2,file=paste0(outDir,"sample_collection_summary.csv"))

```

#PERMANOVA on all samples 
Effects of sponge Order and geography on composition.
```{r adonis, results="markdown"}
load("01_results/psCLR.rds")
load("01_results/DistACH.rds")
#permanova
library(vegan)
samp<-as(sample_data(psCLR),"data.frame")
if(!file.exists(paste0(outDir,"bigAdonis.rds"))){
#ado<-adonis(otu_table(psCLR) ~ phyloOrder+locale,data=samp,permutations = 10)
ado<-adonis(DistACH ~ phyloOrder+locale,data=samp,permutations = 1000)
save(ado,file=paste0(outDir,"bigAdonis.rds"))
}else{load(paste0(outDir,"bigAdonis.rds"))}
ado
```

```{r mds,eval=F}
#Ordination of all samples, colored by clade
#tree with clades annotated
if(!file.exists(paste0(outDir,"ords.Rds"))){
ord.nmds.veg1 <- vegan::metaMDS(DistACH, distance="euclidean",k=2)
#noconvergence
ord.nmds.veg2 <- vegan::metaMDS(DistACH, distance="euclidean",k=3)
#library("plot3D")
# scatter3D(x, y, z, phi = 0, bty = "g", pch = 20, cex = 2, ticktype = "simple")
ord.nmds.veg3 <- vegan::metaMDS(DistACH, distance="bray",k=2)
ord.nmds.veg4 <- vegan::metaMDS(DistACH, distance="bray",k=3)
vegan::stressplot(ord.nmds.veg3)

save(ord.nmds.veg1,ord.nmds.veg2,ord.nmds.veg3,ord.nmds.veg4,file=paste0(outDir,"ords.Rds"))
}else{load(paste0(outDir,"ords.Rds"))}
pMDS<-plot_ordination(psClR_match, ord.nmds.veg, color="phyloOrder",shape="locale")
pMDS
pMDS + facet_grid(~locale)
```

### Figure1. Composition summary PCA
```{r PCA}


 if(!file.exists(paste0(outDir,"logrankPCA.rds"))){
 
  pslog <- transform_sample_counts(ps4, function(x) log(1 + x))
abund <- t(otu_table(pslog))
#hist(sample_sums(ps4),breaks=1000)
#hist(sample_sums(pslog),breaks=1000)
qplot(colSums(otu_table(ps4)),binwidth=1000) +
  xlab("counts-per-sample (raw)")
qplot(log10(colSums(otu_table(ps4))),binwidth=0.01) +
  xlab("counts-per-sample (log10)")
abund_ranks <- t(apply(abund, 1, rank)) #taxa are col now, want to rank within samples
 head(abund_ranks[,1:10])
hist(abund_ranks)

 
 library(reshape2)
abund_df <- melt(abund, value.name = "abund") %>%
  left_join(melt(abund_ranks, value.name = "rank"))
colnames(abund_df) <- c("sample", "seq", "abund", "rank")
 head(abund_df)

 #visualize ranks vs abundance to determine threshold for tying ranks
 sample_ix <- sample(1:nrow(abund_df), 20)
ggplot(abund_df %>%
          filter(sample %in% abund_df$sample[sample_ix])) +
  geom_point(aes(x = abund, y = rank, col = sample),
              position = position_jitter(width = 0.2), size = .7) +
  labs(x = "Abundance", y = "Thresholded rank") 
mean(abund_df$rank)
quantile(abund_df$rank)
#set threshold to minimize effect of rarest ASVs
  abund_ranks <- abund_ranks - mean(abund_ranks)
abund_ranks[abund_ranks < 1] <- 1
hist(abund_ranks[abund_ranks>1])
keep<-colSums(abund_ranks)!=nrow(abund_ranks)
#filter out lowest ranking taxa
pslog2rank<-prune_taxa(keep,pslog)
abund_ranks2<-abund_ranks[,keep ]


library(ade4)
ranks_pca <- dudi.pca(t(abund_ranks2), scannf = F, nf = 3) #transpose to match example code
#set up CCA
ps_ccpna <- ordinate(pslog2rank, "CCA", formula = pslog2rank ~ locale + phyloOrder)
ps_ccpnaGEO <- ordinate(pslog2rank, "CCA", formula = pslog2rank ~ locale )
ps_ccpnaPHYLO <- ordinate(pslog2rank, "CCA", formula = pslog2rank ~ phyloOrder)
save(abund_ranks2,pslog2rank,ranks_pca,ps_ccpna,ps_ccpnaGEO,ps_ccpnaPHYLO,ps_ccpnaPHYLO,file=paste0(outDir,"logrankPCA.rds"))
}else{load(paste0(outDir,"logrankPCA.rds"))}

row_scores <- data.frame(li = ranks_pca$li,
                            seq = colnames(abund_ranks2))
col_scores <- data.frame(co = ranks_pca$co,
                            SampleID = rownames(abund_ranks2))

tax <- tax_table(pslog2rank)@.Data %>%
  data.frame(stringsAsFactors = FALSE)
tax$seq <- rownames(tax)
main_phyla <- c("Chloroflexi", "Proteobacteria", "Cyanobacteria",
                   "PAUC34f","Poribacteria","Acidobacteria")
tax$Phylum[!(tax$Phylum %in% main_phyla)] <- "Other"
tax$Phylum <- factor(tax$Phylum, levels = c(main_phyla, "Other"))
tax$otu_id <- seq_len(nrow(tax_table(pslog2rank)))
row_scores <- row_scores %>% left_join(tax)

samp<-data.frame(sample_data(pslog2rank))
col_scores<-left_join(samp %>%  mutate(SampleID = rownames(samp)),col_scores, by = 'SampleID')
 ordlev<-names(colsC)[names(colsC) %in% col_scores$phyloOrder]
col_scores$phyloOrder<-factor(col_scores$phyloOrder,levels=ordlev)





#PLOT PCA fviz(dudi) too slow?
library(factoextra)
fviz_eig(ranks_pca)


#PLOT PCA ggplot(dudi)
evals_prop <- 100 * (ranks_pca$eig / sum(ranks_pca$eig))
pca.samp<-ggplot() +  labs(x = sprintf("Axis1 [%s%% variance]", round(evals_prop[1], 2)),  y = sprintf("Axis2 [%s%% variance]", round(evals_prop[2], 2)))+ coord_fixed(sqrt(ranks_pca$eig[2] / ranks_pca$eig[1])) +
  geom_point(data = col_scores, aes(x = co.Comp1, y = co.Comp2,shape=locale,color=phyloOrder),size = 1, alpha = 0.8) +
  scale_color_manual(values=colsC) +theme(panel.border = element_rect(color = "#787878", fill = alpha("white", 0))) + guides(col = guide_legend(override.aes = list(size = 3)),shape = guide_legend(override.aes = list(size = 3))) 
pca.samp 
pca.samp +  facet_wrap(~ locale,nrow=2) + theme(legend.position = "none")


pca.mphy<-ggplot() +  labs(x = sprintf("Axis1 [%s%% variance]", round(evals_prop[1], 2)),  y = sprintf("Axis2 [%s%% variance]", round(evals_prop[2], 2))) + coord_fixed(sqrt(ranks_pca$eig[2] / ranks_pca$eig[1])) +  
  geom_point(data = row_scores, aes(x = li.Axis1, y = li.Axis2, col = Phylum))
  pca.mphy
 
   pca.mphy +theme(panel.border = element_rect(color = "#787878", fill = alpha("white", 0))) + guides(col = guide_legend(override.aes = list(size = 3))) 

#PLOT CCA
library(ggrepel)

plotCCAcoords<-function(psccpna){
  ps_scores <- vegan::scores(ps_ccpna)
sites <- data.frame(ps_scores$sites) #samples
sites$SampleID <- rownames(sites)
sites <-left_join(samp %>%  mutate(SampleID = rownames(samp)),sites, by = 'SampleID')

ordlev<-names(colsC)[names(colsC) %in% sites$phyloOrder]
sites$phyloOrder<-factor(sites$phyloOrder,levels=ordlev)
sites$locale
species <- data.frame(ps_scores$species) #ASV variable
species <- species %>%  mutate(seq = rownames(species))
  species<-left_join(tax %>%  mutate(seq = rownames(tax)),species ,by='seq')
evals_prop <- 100 * ps_ccpna$CCA$eig[1:2] / sum(ps_ccpna$CA$eig)
ccpnaplot<- ggplot()  + guides(col = guide_legend(override.aes = list(size = 3))) + coord_fixed(sqrt(ps_ccpna$CCA$eig[2] / ps_ccpna$CCA$eig[1])*0.33) +
  labs(x = sprintf("Axis1 [%s%% variance]", round(evals_prop[1], 2)), 
       y = sprintf("Axis2 [%s%% variance]", round(evals_prop[2], 2))) +  theme(panel.border = element_rect(color = "#787878", fill = alpha("white", 0))) +
  geom_point(data = sites, aes(x = CCA1, y = CCA2,color=phyloOrder,shape = locale), alpha = 0.5) +  scale_color_manual(values=colsC[ordlev])  
ccpnaplot
ccpnaplot +  facet_wrap(~ locale,nrow=2) + theme(legend.position = "none")
#ccpnaplot + geom_point(data = species, aes(x = CCA1, y = CCA2), size = 0.5) + geom_text_repel(data = species %>% filter(CCA2 < -2), aes(x = CCA1, y = CCA2, label = Phylum),   size = 1.5, segment.size = 0.1) 

}
cat("CCA: log-counts ~ location + host order")
plotCCAcoords(ps_ccpna)
cat("CCA: log-counts ~ location")
plotCCAcoords(ps_ccpnaGEO)
cat("CCA: log-counts ~  host order")
plotCCAcoords(ps_ccpnaPHYLO)




```

##Bray distance network between samples
```{r network,results="markdown"}
set.seed(711L)
#physeqs method excludes most samples...try igraph network with ggnet2 plotting  https://briatte.github.io/ggnet/
library(igraph)
library(GGally)
net5 <- make_network(pslog2rank, max.dist=0.5)
net75 <- make_network(pslog2rank, max.dist=0.75)
net8 <- make_network(pslog2rank, max.dist=0.8)
net85 <- make_network(pslog2rank, max.dist=0.85)
net9 <- make_network(pslog2rank, max.dist=0.9)
net95 <- make_network(pslog2rank, max.dist=0.95)

samp<-data.frame(sample_data(pslog2rank))
ordlev<-names(colsC)[names(colsC) %in% samp$phyloOrder]
samp$phyloOrder<-factor(samp$phyloOrder,levels=ordlev)

net_prep<-function(net){
V(net)$locale <- samp[names(V(net)), "locale"]
V(net)$locale<-levels(samp$locale)[V(net)$locale]
V(net)$phyloOrder<- samp[names(V(net)), "phyloOrder"]
V(net)$phyloOrder<-levels(samp$phyloOrder)[V(net)$phyloOrder]
print(paste0("Number of samples retained: ",length(names(V(net)))," out of ",length(samp$DNAid)))
print(ggnet2(net,color="phyloOrder",shape="locale",node.size=1,palette=colsC[ordlev]))
}
cat("Max distance 0.5")
net<-net_prep(net5)
cat("Max distance 0.75")
net<-net_prep(net75)
cat("Max distance 0.8")
net<-net_prep(net8)
cat("Max distance 0.85")
net<-net_prep(net85)
cat("Max distance 0.9")
net<-net_prep(net9)
cat("Max distance 0.95")
net<-net_prep(net95)

#pg<-ggnet2(net,color="phyloOrder",shape="locale",node.size=1,palette=colsC[ordlev]);pg

```

## Minimum Spanning Tree with Jaccard distance
Testing maximum distances of 0.75 to 0.95
```{r mst,results="markdown"}
pdf(paste0(outDir,"fig1_mst.pdf"),height=8,width = 8)
library(phyloseqGraphTest)
###test first
#ps<-psCLR
ps<-ps4
samp<-data.frame(sample_data(ps))
ordlev<-names(colsC)[names(colsC) %in% samp$phyloOrder]
samp$phyloOrder<-factor(samp$phyloOrder,levels=ordlev)

 gt <- graph_perm_test(ps, "phyloOrder", 
                      distance = "bray", type = "mst",max.dist=0.5)
gt$sampletype<-factor(gt$sampletype,levels=levels(samp$phyloOrder))
ordlev<-names(colsC)[names(colsC) %in% samp$phyloOrder]

plotNet1=plot_test_network(gt) + theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))  +  scale_color_manual(values=colsC[ordlev]) 
pdf(paste0(outDir,"mstNEW2.pdf"),useDingbats=F)
plotNet1
dev.off()

###
mst_perm_test<-function(mdist){
  gt <- graph_perm_test(ps, "phyloOrder", 
                      distance = "jaccard", type = "mst",max.dist=mdist)

gt$sampletype<-factor(gt$sampletype,levels=levels(samp$phyloOrder))
ordlev<-names(colsC)[names(colsC) %in% samp$phyloOrder]

plotNet1=plot_test_network(gt) + theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))  +  scale_color_manual(values=colsC[ordlev]) 

plotPerm1=plot_permutations(gt)
print(plotNet1)
print(gt$pval)
print(plotPerm1)
}


cat("Max distance 0.5")
mst_perm_test(mdist=0.5)
dev.off()
cat("Max distance 0.75")
mst_perm_test(0.75)
cat("Max distance 0.8")
mst_perm_test(0.8)
cat("Max distance 0.85")
mst_perm_test(0.85)
cat("Max distance 0.9")
mst_perm_test(0.9)
cat("Max distance 0.95")
mst_perm_test(0.95)

```



##Barplot alternative to superimpose over map
```{r maps}
#samp add for each site lat long 


#world <- map_data('world')
#p <- ggplot(world, aes(long, lat)) +
   # geom_map(map=world, aes(map_id=region), fill=NA, color="black") +
  #  coord_quickmap()
#library(treemapify) #update R to use
#p4 +geom_treemap()
spcolors <- setNames(  spbars2,spbars1$Species_consensus)

library(treemap)


samp<-data.frame(sample_data(ps4))
ordlev<-names(colsC)[names(colsC) %in% samp$phyloOrder]
samp$phyloOrder<-factor(samp$phyloOrder,levels=ordlev)

test<-samp %>% group_by(phyloOrder) %>%
  count(Species_consensus)
mapcols<-colsC[names(colsC) %in% test$phyloOrder]

tm1<-treemap(dtf = test,
        index = c("phyloOrder", "Species_consensus"),
        vSize = "n",vColor = "phyloOrder",
        palette=mapcols,type="categorical",
        fontsize.labels = c(10,5))
sx<-samp[samp$locale=="SX",] %>% group_by(phyloOrder) %>% count(Species_consensus)
sxcols<-colsC[names(colsC) %in% sx$phyloOrder]

ky<-samp[samp$locale=="KY",] %>% group_by(phyloOrder) %>% count(Species_consensus)
kycols<-colsC[names(colsC) %in% ky$phyloOrder]

cu<-samp[samp$locale=="CU",] %>% group_by(phyloOrder) %>% count(Species_consensus)
cucols<-colsC[names(colsC) %in% cu$phyloOrder]

bz<-samp[samp$locale=="BZ",] %>% group_by(phyloOrder) %>% count(Species_consensus)
bzcols<-colsC[names(colsC) %in% bz$phyloOrder]


tmsx<-sx %>% treemap(dtf = ., index = c("phyloOrder", "Species_consensus"),
        vSize = "n",vColor = "phyloOrder",palette=sxcols, type="categorical",
        fontsize.labels = c(0,0),title="St Croix")
tmky<-ky %>% treemap(dtf = ., index = c("phyloOrder", "Species_consensus"),
        vSize = "n",vColor = "phyloOrder",palette=kycols, type="categorical",
        fontsize.labels = c(0,0),title="Cayman")
tmcu<-cu %>% treemap(dtf = ., index = c("phyloOrder", "Species_consensus"),
        vSize = "n",vColor = "phyloOrder",palette=cucols, type="categorical",
        fontsize.labels = c(0,0),title="Curacao")
tmbz<-bz %>% treemap(dtf = ., index = c("phyloOrder", "Species_consensus"),
        vSize = "n",vColor = "phyloOrder",palette=bzcols, type="categorical",
        fontsize.labels = c(0,0),title="Belize")






```




