---
title: "Random Forest modelling of MA status"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    toc_depth: 4
    df_print: paged
author: "Sabrina Pankey"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE,fig.width = 8, results="markdown")
options(scipen = 999)
```


```{r load, echo=FALSE, include=FALSE} 
mainDir<-"~/Box Sync/MPL/Lesser DOB sponges/coevolution_analyses/pipeline4/"
setwd(mainDir)
outDir<-"03_char_results/"
ifelse(!dir.exists(file.path(mainDir, outDir)), dir.create(file.path(mainDir, outDir)), FALSE)
library(dplyr)
library(phyloseq)
library(ggplot2)
library(colorspace)

#library(scales);show_col(rainbow_hcl(12))
hma.col<-rainbow_hcl(12)[2]; lma.col<-rainbow_hcl(12)[7]



#read in character data to designate species for training models
chardf<-read.csv("02_input_spongetree/chars_by_species.csv",header = T,stringsAsFactors = F)
char<-chardf %>% select(Species_consensus,consensusmod)
char_trainer<-char[!is.na(char$consensusmod),] ##only species with known status included for training model

load("01_results/psRm_bySpecies.rds")
```

## Random Forest classification of microbiomes into HMA or LMA
#### RF classification using all `r ntaxa(psRm)` ASV abundances

```{r, RandomForestWrapper}
# random forest vs glm
# http://rpubs.com/EAVWing/trees_and_glm
library(randomForest)
library(caret)

runRF<-function(traindf, predictdf){
 rf <- randomForest(consensusmod ~ ., data=traindf, importance=TRUE, proximity=TRUE)
 imps<-importance(rf)
#predict new samples without cross-validation
preds<-predict(rf,predictdf)
rf.mds<-cmdscale(1-rf$proximity,eig=T)
return(list(preds,imps,rf.mds))
}

cross.validateRF<-function(traindf,predictdf){
  rf_caret <- train(form = consensusmod ~ .,
                      data = traindf, method = "rf", importance = TRUE,
                     tuneGrid=expand.grid(.mtry=c(2:round(ncol(traindf)/5),round(ncol(traindf)/4), round(ncol(traindf)/3), round(ncol(traindf)/2),ncol(traindf)-1)),
                      trControl = trainControl(method = "repeatedcv",
                                               number = 10,
                                               repeats = 8))
cvImp<-varImp(rf_caret)
predCV<-predict(rf_caret,predictdf)
#names(allasvRFpredCV)<-rownames(counts)[!(rownames(counts) %in% counts_char_trainer$Species_consensus)]
names(predCV)<-rownames(predictdf)
return(list(predCV,rf_caret,cvImp))
}

cross.validateRF.faster<-function(traindf,predictdf){
  rf_caret <- train(form = consensusmod ~ .,
                      data = traindf, method = "rf", importance = TRUE,
                      trControl = trainControl(method = "repeatedcv",
                                               number = 10,
                                               repeats = 8))
cvImp<-varImp(rf_caret)
predCV<-predict(rf_caret,predictdf)
#names(allasvRFpredCV)<-rownames(counts)[!(rownames(counts) %in% counts_char_trainer$Species_consensus)]
names(predCV)<-rownames(predictdf)
return(list(predCV,rf_caret,cvImp))
}
```

```{r fxn_plotImportance}
#takes rf$importance df as input
msp.varImpPlot<-function (imp, sort = TRUE, n.var = min(30, nrow(imp)), 
    type = NULL, class = NULL, scale = TRUE,  main="Importance",...) {
    #if (!inherits(x, "randomForest")) 
   #     stop("This function only works for objects of class `randomForest'")
  #  imp <- importance(x, class = class, scale = scale, type = type, ...)
    if (ncol(imp) > 2) 
        imp <- imp[, -(1:(ncol(imp) - 2))]
    nmeas <- ncol(imp)
    if (nmeas > 1) {
        op <- par(mfrow = c(1, 2), mar = c(4, 5, 4, 1), mgp = c(2, 
            0.8, 0), oma = c(0, 0, 2, 0), no.readonly = TRUE)
        on.exit(par(op))
    }
    for (i in 1:nmeas) {
        ord <- if (sort) 
            rev(order(imp[, i], decreasing = TRUE)[1:n.var])
        else 1:n.var
        xmin <- if (colnames(imp)[i] %in% c("IncNodePurity", 
            "MeanDecreaseGini")) 
            0
        else min(imp[ord, i])
        dotchart(imp[ord, i], xlab = colnames(imp)[i], ylab = "",cex=.5, 
            main = if (nmeas == 1) 
                main
            else NULL, xlim = c(xmin, max(imp[, i])), ...)
    }
    if (nmeas > 1) 
        mtext(outer = TRUE, side = 3, text = main, cex = 1.2)
    invisible(imp)
}
```

```{r runRF_all,results="markdown"}

#convert to relative abundance since species have different read totals
relabund.spp<- transform_sample_counts(psRm, function(OTU) OTU/sum(OTU)) #convert to relative abundance
#sample_sums(relabund.spp) #should all be 1
#min(taxa_sums(relabund.spp))
relabund.spp1<-prune_taxa(taxa_sums(relabund.spp)>0.001,relabund.spp)

if(!(file.exists("RFallASV.rds"))){
 
counts<-round(data.frame(otu_table(relabund.spp1)),digits = 4)#need to round becuase hitting memory errors in RF execution
counts$Species_consensus<-rownames(counts)
 

counts_char_trainer<-inner_join(char_trainer,counts, by="Species_consensus")
rownames(counts_char_trainer)<-counts_char_trainer[,1]
data2train<-counts_char_trainer[,-1] #drop Species_consensus column
data2train$consensusmod<-as.factor(data2train$consensusmod)

data2predict<-counts[!(rownames(counts) %in% counts_char_trainer$Species_consensus),]
  
start1<-Sys.time()
set.seed(33)
#run RF functions
out1<-runRF(data2train,data2predict)
all_predstates<-out1[[1]]
imp.all<-out1[[2]]
all.mds<-out1[[3]]
cat ("GOF:");print(all.mds$GOF)
plot(all.mds$points[,1],all.mds$points[,2])
out2<-cross.validateRF.faster(data2train,data2predict)
all_predstatesCV<-out2[[1]]; RFcv<-out2[[2]]; 
print(RFcv)

end1<-Sys.time()
cat("Random Forest and Cross-validation of HMA/LMA states using all ASVs took:")
print(end1-start1) #Time difference of 46 mins

save(all_predstates,all_predstatesCV,out1,file=paste0(outDir,"RFallASV.rds"))
}else{load(paste0(outDir,"RFallASV.rds"))}
#compare original predictions to cross-validated prediction

cbind(all_predstates,all_predstatesCV)
all_predstates<-out1[[1]]
imp.all<-out1[[2]]
tmp<-as.data.frame(imp.all)
#tmp[rev(order(tmp$MeanDecreaseAccuracy)),]

all.mds<-out1[[3]]
cat("GOF:"); print(all.mds$GOF)
plot(all.mds$points[,1],all.mds$points[,2])

topimps<-tmp[which(tmp$MeanDecreaseAccuracy>0),]
#hist(topimps$MeanDecreaseAccuracy)
#quantile(topimps$MeanDecreaseAccuracy)
tax_table(relabund.spp1)[rownames(topimps),"KPC"]

rownames(imp.all)<-c(tax_table(relabund.spp1)[rownames(tax_table(relabund.spp1)) %in% rownames(imp.all),"KPC"])
msp.varImpPlot(imp.all)

topimps$KPC<-c(tax_table(relabund.spp1)[rownames(tax_table(relabund.spp1)) %in% rownames(topimps),"KPC"])
#table(topimps$KPC)
barplot(table(topimps$KPC),cex.names=0.4,horiz=T,las=2,xlab="RF Importance score (MeanDecreaseAccuracy)")

#save predictions later, after adding a column in imp.phy ascribing which HMA LMA condition is more enriched for each phylum
#save(imp.phy,file="RFimportance_phyla.Rds")

library(tidyr)
topimps2<-imp.all[rowSums(imp.all[,1:2])>median(rowSums(imp.all[,1:2])),]
topimps3<-topimps2[rowSums(topimps2[,1:2])>median(rowSums(topimps2[,1:2])),]

imp2<-data.frame(topimps3[rev(order(rowSums(topimps3[,1:2]))),1:2]); imp2$KP<-rownames(imp2)
gathered<-gather(imp2,variable,value, -KP)
gathered$KP<-factor(gathered$KP,levels=imp2$KP)

ggplot(gathered, aes(x=value, y=KP)) +   
  geom_bar(aes(fill = variable), position = "dodge", stat="identity") +
  theme(axis.text.y=element_text(angle=0,hjust=1,vjust=0.5,size=2))

```

#### Predict HMA-LMA status using overall abundance of microbial PHYLA only
 
```{r RFphyla}
phylumGlommed = tax_glom(psRm, taxrank="Phylum")
tax_table(phylumGlommed)[,"KP"]<-paste0(tax_table(phylumGlommed)[,"Kingdom"],"_",tax_table(phylumGlommed)[,"Phylum"])
#convert to relative abundance since species have different read totals
relabund.phy<- transform_sample_counts(phylumGlommed, function(OTU) OTU/sum(OTU)) 
save(phylumGlommed,relabund.phy,file=paste0(outDir,"psRm.phylum.Rds"))
#convert to relative abundance
sample_sums(relabund.phy) #should all be 1

phycounts1<-as.data.frame(otu_table(relabund.phy))
phycounts2<-round(phycounts1,digits=4) #need to round becuase hitting memory errors in RF execution
phycounts3<-phycounts2[,colSums(phycounts2)>0]

tmp<-round(otu_table(relabund.phy),digits = 4)#need to round becuase hitting memory errors in RF execution
relabund.phy1<-prune_taxa(taxa_sums(tmp)>0.001,tmp)
min(taxa_sums(relabund.phy1))
phycounts1<-data.frame(relabund.phy1)
ncol(phycounts1) 
phycounts1<-phycounts1[,colSums(phycounts1)>0]


phycounts<-phycounts1[rownames(phycounts1) %in% char_trainer$Species_consensus,]
colSums(phycounts)
phycounts$Species_consensus<-rownames(phycounts)
phycounts_char_trainer<-inner_join(char_trainer,phycounts, by="Species_consensus")

rownames(phycounts_char_trainer)<-phycounts_char_trainer[,1]
data2train<-phycounts_char_trainer[,-1]
data2train$consensusmod<-as.factor(data2train$consensusmod)

#setup data to be fit using phyla abundances
data2predict<-phycounts3[!(rownames(phycounts3) %in% phycounts_char_trainer$Species_consensus),]

```
RF classification using only agglomerated abundances of `r ntaxa(phylumGlommed)` phyla detected across `r nsamples(psRm)` species.

```{r RFphyla2,results="markdown"}


#run RF functions
out<-runRF(data2train,data2predict)
phy_predstates<-out[[1]]
imp.phy<-out[[2]]
phy.mds<-out[[3]]
print(phy.mds$GOF)
plot(phy.mds$points[,1],phy.mds$points[,2])
out<-cross.validateRF(data2train,data2predict)
phy_predstatesCV<-out[[1]]; phyRFcv<-out[[2]]; 
print(phyRFcv)
#phyImportance<-out[[3]]
#plot(phyImportance) #use RF imp measures for plotting

#compare:
cbind(all_predstatesCV,phy_predstates,phy_predstatesCV)


```

Importance measures for phyla in RF classification. 

```{r}
rownames(imp.phy)<-c(tax_table(phylumGlommed)[rownames(tax_table(phylumGlommed)) %in% rownames(imp.phy),"KP"])

msp.varImpPlot(imp.phy)

#save predictions later, after adding a column in imp.phy ascribing which HMA LMA condition is more enriched for each phylum
#save(imp.phy,file="RFimportance_phyla.Rds")

library(tidyr)
imp2<-data.frame(imp.phy[rev(order(rowSums(imp.phy[,1:2]))),1:2]); imp2$KP<-rownames(imp2)
gathered<-gather(imp2,variable,value, -KP)
gathered$KP<-factor(gathered$KP,levels=imp2$KP)

ggplot(gathered, aes(x=value, y=KP)) +   
  geom_bar(aes(fill = variable), position = "dodge", stat="identity") +
  theme(axis.text.x=element_text(angle=0,hjust=1,vjust=0.5,size=3))

```


#### Predict HMA-LMA status using abundance of microbial CLASSES only
 
```{r RFclass1}
classGlommed = tax_glom(psRm, taxrank="Class")
tax_table(classGlommed)[,"KPC"]<-paste0(tax_table(classGlommed)[,"Kingdom"],"_",tax_table(classGlommed)[,"Phylum"],"_",tax_table(classGlommed)[,"Class"])
#convert to relative abundance since species have different read totals
relabund.class<- transform_sample_counts(classGlommed, function(OTU) OTU/sum(OTU)) #convert to relative abundance
sample_sums(relabund.class) #should all be 1
save(classGlommed,relabund.class,file=paste0(outDir,"psRm.class.Rds"))

clcounts1<-as.data.frame(otu_table(relabund.class))
clcounts2<-round(clcounts1,digits=4) #need to round becuase hitting memory errors in RF execution
colSums(clcounts2)
clcounts3<-clcounts2[,colSums(clcounts2)>0]
clcounts3$Species_consensus<-rownames(clcounts3)

clcounts<-clcounts3[rownames(clcounts3) %in% char_trainer$Species_consensus,]
clcounts_char_trainer<-inner_join(char_trainer,clcounts, by="Species_consensus")
rownames(clcounts_char_trainer)<-clcounts_char_trainer[,1]
data2train<-clcounts_char_trainer[,-1]

data2train$consensusmod<-as.factor(data2train$consensusmod)

#setup data to be fit using phyla abundances
data2predict<-clcounts3[!(rownames(clcounts3) %in% phycounts_char_trainer$Species_consensus),]


```
RF classification using only agglomerated abundances of `r ntaxa(classGlommed)` classes detected across `r nsamples(psRm)` species.

```{r RFclass2,results="markdown"}
out<-runRF(data2train,data2predict)
cl_predstates<-out[[1]]
imp.cl<-out[[2]]
cl.mds<-out[[3]]
print(cl.mds$GOF)
plot(cl.mds$points[,1],cl.mds$points[,2])
out<-cross.validateRF(data2train,data2predict)
cl_predstatesCV<-out[[1]]; clRFcv<-out[[2]]; 
print(clRFcv)

rownames(imp.cl)<-c(tax_table(classGlommed)[rownames(tax_table(classGlommed)) %in% rownames(imp.cl),"KPC"])

msp.varImpPlot(imp.cl)

library(tidyr)
imp2<-data.frame(imp.cl[rev(order(rowSums(imp.cl[,1:2]))),1:2]); imp2$KP<-rownames(imp2)

gathered<-gather(imp2,variable,value, -KP)
gathered$KP<-factor(gathered$KP,levels=imp2$KP)

ggplot(gathered, aes(x=value, y=KP)) +   
  geom_bar(aes(fill = variable), position = "dodge", stat="identity") +
  theme(axis.text.x=element_text(angle=0,hjust=1,vjust=0.5,size=3))


#compare:
cbind(all_predstatesCV,phy_predstatesCV,cl_predstatesCV)
```

Comparison of sponges species' predictions that differ depending on level used during RF classification.
```{r,results="markdown"}
#add predictions to character dataframe
tmp<-char; rownames(tmp)<-tmp$Species_consensus

pred<-as.character(all_predstatesCV); names(pred)<-names(all_predstatesCV)
char$consensusRF_asv<-ifelse(is.na(tmp$consensusmod), pred[rownames(tmp)], tmp$consensusmod)

pred2<-as.character(phy_predstatesCV); names(pred2)<-names(phy_predstatesCV)
char$consensusRF_phyla<-ifelse(is.na(tmp$consensusmod), pred2[rownames(tmp)], tmp$consensusmod)

pred3<-as.character(cl_predstatesCV); names(pred3)<-names(cl_predstatesCV)
char$consensusRF_class<-ifelse(is.na(tmp$consensusmod), pred2[rownames(tmp)], tmp$consensusmod)

save(char,file=paste0(outDir,"species_charsRF.Rds"))
char %>% filter(consensusRF_phyla!= consensusRF_asv |consensusRF_phyla!= consensusRF_class | consensusRF_class!= consensusRF_asv)
```



```{r saveImportanceRFpreds}

#save Importance values of each Phylum, 
#add in columns wtih 
# -mean relative abundance of phylum in each type LMAra HMAra
# -pval for differential abundance pval
# -which type is enriched enrichedIn

load("01_results/psR.rds")
sdat<-data.frame(sample_data(psR)); sdat$X<-rownames(sdat)
sdat<-sdat %>% left_join(.,char); rownames(sdat)<-sdat$X
sample_data(psR)<-sdat

phylumGlommed<-tax_glom(psR,taxrank="Phylum")
tax_table(phylumGlommed)[,"KP"]<-paste0(tax_table(phylumGlommed)[,"Kingdom"],"_",tax_table(phylumGlommed)[,"Phylum"])
keep<-sample_names(phylumGlommed)[!is.na(sample_data(phylumGlommed)$consensusRF_phyla)]
p2<-prune_samples(keep,phylumGlommed)

ps2df<-function(ps){
OTU1 = as(otu_table(ps), "matrix")
# transpose if necessary
if(taxa_are_rows(ps)){OTU1 <- t(OTU1)}
# Coerce to data.frame
OTUdf = as.data.frame(OTU1)
return(OTUdf)
}
psdf<-ps2df(p2)
phyla_pvals<-apply(psdf,2,function(p) 
  {ttest<-t.test(p~factor(sample_data(p2)$consensusRF_phyla)) 
  return(ttest$p.value)})
names(phyla_pvals)<-tax_table(p2)[,"KP"]

#merge_samples sums otu_counts, instead of averaging across sample factor.  Manually calc average for HMA and LMA
PGroup <- transform_sample_counts(p2, function(x) x / sum(x))
SampleType <- factor(sample_data(PGroup)$consensusRF_phyla)
table(SampleType)

mean_PGroup <- sapply(levels(SampleType),function(i){
  rowMeans(otu_table(PGroup)[,SampleType==i])
})

phy <- tax_table(PGroup)[rownames(mean_PGroup ),"KP"]
rownames(mean_PGroup) <- phy
colnames(mean_PGroup)<-c("HMAra","LMAra")
mean_PGroup<-data.frame(mean_PGroup)
mean_PGroup$bacterial_group<-rownames(mean_PGroup)

#combine results, add to dataframe for plotting
enrichedP<-round(phyla_pvals,8) %>% data.frame; colnames(enrichedP)<-"pval"
enrichedP$bacterial_group<-rownames(enrichedP)


imp.phy2<-data.frame(imp.phy)
imp.phy2$bacterial_group<-rownames(imp.phy2)
imp.phy2<-left_join(imp.phy2,mean_PGroup) %>% left_join(.,enrichedP)
imp.phy2<-imp.phy2 %>% mutate(HMLvLMA=ifelse(HMAra>LMAra,HMAra*100,(-100*LMAra)))

imp.phy2<-imp.phy2 %>% mutate(enrichedIn=ifelse(HMAra>LMAra,"HMA","LMA"))
imp.phy<-imp.phy2
save(imp.phy,file=paste0(outDir,"RFimportance_phyla.Rds"))





cGlommed<-tax_glom(psR,taxrank="Class")
tax_table(cGlommed)[,"KPC"]<-paste0(tax_table(cGlommed)[,"Kingdom"],"_",tax_table(cGlommed)[,"Phylum"],"_",tax_table(cGlommed)[,"Class"])
keep<-sample_names(cGlommed)[!is.na(sample_data(cGlommed)$consensusRF_phyla)]
p3<-prune_samples(keep,cGlommed)
psdf<-ps2df(p3)
class_pvals<-apply(psdf,2,function(p) 
  {ttest<-t.test(p~factor(sample_data(p3)$consensusRF_phyla)) 
  return(ttest$p.value)})
names(class_pvals)<-tax_table(p3)[,"KPC"]

CGroup<-transform_sample_counts(p3,fun= function(otu) otu/sum(otu))
SampleType <- factor(sample_data(CGroup)$consensusRF_phyla)
table(SampleType)

mean_CGroup <- sapply(levels(SampleType),function(i){
  rowMeans(otu_table(CGroup)[,SampleType==i])
})

clas <- tax_table(p3)[rownames(mean_CGroup ),"KPC"]
rownames(mean_CGroup) <- clas

colnames(mean_CGroup)<-c("HMAra","LMAra")
mean_CGroup<-data.frame(mean_CGroup)
mean_CGroup$bacterial_group<-rownames(mean_CGroup)

#combine results, add to dataframe for plotting
enrichedP<-round(class_pvals,8) %>% data.frame; colnames(enrichedP)<-"pval"
enrichedP$bacterial_group<-rownames(enrichedP)


imp.cl2<-data.frame(imp.cl)
imp.cl2$bacterial_group<-rownames(imp.cl2)
imp.cl2<-left_join(imp.cl2,mean_CGroup) %>% left_join(.,enrichedP)
imp.cl2<-imp.cl2 %>% mutate(HMLvLMA=ifelse(HMAra>LMAra,HMAra*100,(-100*LMAra)))

imp.cl2<-imp.cl2 %>% mutate(enrichedIn=ifelse(HMAra>LMAra,"HMA","LMA"))
imp.cl<-imp.cl2
save(imp.cl,file=paste0(outDir,"RFimportance_class.Rds"))

```

```{r setIimportanceThreshold}

modImps<-function(imps){
  #eliminate less/non predictive taxa
  imps<-imps[!(is.na(imps$enrichedIn)),] #NA = goups where avg rel abund is 0 for both
  #order: HMA (strongest to weakest) then LMA (weakest to strongest)
  imps<-imps[imps$MeanDecreaseAccuracy>0,]
  #ord:
  negAcc<-imps$MeanDecreaseAccuracy*-1
  imps<-cbind.data.frame(imps,negAcc)
  AccOrder<-apply(imps,1, function(x) ifelse(x["enrichedIn"]=="LMA",x["negAcc"],x["MeanDecreaseAccuracy"]))
  imps<-cbind.data.frame(imps,AccOrder,stringsAsFactors=F)
  imps<-imps[rev(order(as.numeric(imps$AccOrder))),]
}
imp.phy<-modImps(imp.phy)
imp.cl<-modImps(imp.cl)
save(imp.phy,imp.cl,file=paste0(outDir,"mostImp.Rds"))
```
