---
title: "Ancestral state recon"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    toc_depth: 4
    df_print: paged
author: "Sabrina Pankey"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE,results="hide",fig.width = 8,fig.height=12)
```


```{r load}
mainDir<-"~/Box Sync/MPL/Lesser DOB sponges/coevolution_analyses/pipeline4/"
setwd(mainDir)
outDir<-"05_calib_results/"
ifelse(!dir.exists(file.path(mainDir, outDir)), dir.create(file.path(mainDir, outDir)), FALSE)
library(ape)
library(ggtree)
library(ggplot2)
library(phytools)
library(strap)
library(diversitree)
library(colorspace)
hma.col<-rainbow_hcl(12)[2]; lma.col<-rainbow_hcl(12)[7]
```

## ASR for HMA,LMA states on species phylogeny 

- time-calibrated chronogram, from TreePL dating

Tips without microbial states known or predicted are removed, since the likelihood models perform perfom poorly with missing data (diversitree documentation).

## Lineage through time on full time tree
```{r loadtimetree}

### load previous time-calib tree...
load("02_tree_results/spongespeciesTimeTree.OGless.Rds") #loads 'timetr', treemeta.u, cladelist&cladecols (==phylo clade colors),colsO & colsF (==WPD taxonomy colors)
ctr<-timetr
any(ctr$edge.length<=0)
min(ctr$edge.length[ctr$edge.length>0])
ctr$edge.length[ctr$edge.length==0]<-0.000001
any(ctr$edge.length<=0)
dn<-dist.nodes(ctr)
ctr$root.time<-max(dn)/2
library(strap)
pdf(paste0(outDir,"geoscale_ctrwhole.pdf"),height=11)
geoscalePhylo(tree=ctr, units=c("Period"), boxes="Period")
dev.off()

```

Calibrated tree contains `r Ntip(ctr)` tips.

```{r plotLTT,results="markdown",fig.height=6}
 ltt<-ltt(ctr, log.lineages=F)
cat(paste0("Pybus & Harvey gamma on big timetree: ",ltt$gamma,", p= ",ltt$p))
#TODO: replace with post dist or bootstraps to run CI on lineage #s thru time
#obj<-ltt95(trees,log=F)
#plot(obj,xaxis="flipped")

```

```{r loadCharStates}
load("03_char_results/species_charsRF.Rds") #loads 'char' with consensusRF column containing Random-Forest microbe state predictions
#select which RF prediction to use (based on ASVs, phylum or class)
consensusRF<-char$consensusRF_class

#binarize
char01<-consensusRF;names(char01)<-char$Species_consensus
char01[char01=="HMA"]<-1; char01[char01=="LMA"]<-0; 
char01<-setNames(as.numeric(char01), names(char01))

char01<-char01[names(char01) %in% ctr$tip.label]
char01<-char01[ctr$tip.label]


#retain only species with nonNA characters ...Diversitree MLE performs poorly
todrop<-names(char01)[is.na(char01)]
char01.complete<-char01[!(is.na(char01))]

tokeep<-ctr$tip.label[(ctr$tip.label %in% names(char01.complete))]
ctr.complete<-keep.tip(ctr,tip=tokeep)

#drop any char entries not in tree
char01.complete<-char01.complete[names(char01.complete) %in% ctr.complete$tip.label]
length(char01.complete)==length(ctr.complete$tip.label)
char01.complete<-char01.complete[ctr.complete$tip.label]



tp<-plot(ctr.complete,cex=0.3,no.margin=F,main="Calibrated tree with states")
axisPhylo()
tiplabels(tip=which(ctr.complete$tip.label %in%   names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")

dn<-dist.nodes(ctr.complete)
ctr.complete$root.time<-max(dn)/2
library(strap)
pdf(paste0(outDir,"geoscale_ctrcomplete.pdf"),height=11)
geoscalePhylo(tree=ctr.complete, units=c("Period"), boxes="Period")
dev.off()
#libary(MCMCtreeR)
#MCMC.tree.plot(phy, cex.tips = 0.2, time.correction = 100, scale.res = c("Eon", 
 #   "Period"), plot.type = "phylogram", cex.age = 0.6, cex.labels = 0.6, 
 #   relative.height = 0.08, col.tree = "grey40", label.offset = 4, 
 #   node.method = "none", no.margin = TRUE)
```

Calibrated tree contains `r Ntip(ctr.complete)` tips with HMA/LMA states.

```{r aceASR}

any(ctr.complete$edge.length<=0)


fit.ace.er2<-ace(as.factor(char01.complete),ctr.complete,model="ER", type="discrete",marginal=F)
fit.ace.ard2<-ace(as.factor(char01.complete),ctr.complete,model="ARD", type="discrete",marginal=F)
# 
ifelse(AIC(fit.ace.ard2)<AIC(fit.ace.er2),bestacemodel<-fit.ace.ard2,bestacemodel<-fit.ace.er2)
aic<-as.matrix(sapply(list(ER=fit.ace.er2,ARD=fit.ace.ard2),AIC))
print(aic)

plot(ctr.complete,cex=0.3,no.margin=F,main="Calibrated tree with ace ASR")
axisPhylo()
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
op <- par(fg = "transparent")
nodelabels(pie=bestacemodel$lik.anc,cex=0.5,piecol=c(lma.col,hma.col),cex=0.5)

#nodelabels(pch=16,col="red",node = which(nodeMLstate==0)+Ntip(ctr.complete))
```

## ACE ASR on full tree (many tips missing character data)
```{r incompletedata}
char.mat<-data.frame(to.matrix(char01,c("0","1")))
char.mat[which(rowSums(char.mat)==0),1:2]<-c(NA,NA)
names(char.mat)<-c("LMA","HMA")

dat2<-data.frame(char01)
names(dat2)<-"microbialState"
diversitree::trait.plot(ctr,type="f",dat=dat2,cols=list(microbialState=c(lma.col,hma.col)),lab=c("LMA","HMA"),cex.lab=0.1,legend=T, edge.width=0.1)
#todo plot trait legend
#

#try on calibrated, full tree
fit.ace.er1<-ace(char01,ctr,model="ER", type="discrete",marginal=F)
fit.ace.ard1<-ace(char01,ctr,model="ARD", type="discrete",marginal=F)
# 
ifelse(AIC(fit.ace.ard1)<AIC(fit.ace.er1),bestacemodel1<-fit.ace.ard1,bestacemodel1<-fit.ace.er1)
aic<-as.matrix(sapply(list(ER=fit.ace.er1,ARD=fit.ace.ard1),AIC))
print(aic)

plot(ctr,"fan",cex=0.2,no.margin=F,main="Calibrated tree with ace ASR")
tiplabels(tip=which(ctr$tip.label %in% names(char01)[char01==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr$tip.label %in% names(char01)[char01==1]),col=hma.col,pch=16,cex=1,frame="none")
op <- par(fg = "transparent")
nodelabels(pie=bestacemodel1$lik.anc,piecol=c(lma.col,hma.col),cex=0.3)


plot(ctr,cex=0.3,no.margin=F,main="Calibrated tree with ace ASR")
axisPhylo()
tiplabels(tip=which(ctr$tip.label %in% names(char01)[char01==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr$tip.label %in% names(char01)[char01==1]),col=hma.col,pch=16,cex=1,frame="none")
nodelabels(pie=bestacemodel1$lik.anc,cex=0.5,piecol=c(lma.col,hma.col),cex=0.5)
#

```

### Trait thru time on ACE ASRs
```{r tttfunc,results="markdown",fig.height=6}
#testing
#phy<-ctr.complete; ASR01<-nodeMLstate; tipchar01<-char01.complete;colors01<-c(lma.col,hma.col)


plot_traits_thru_time<-function(phy,ASR01,tipchar01,colors01){

edges<-data.frame(phy$edge);colnames(edges)<-c("anc","desc")

parentstates<-cbind(ASR01,(Ntip(phy)+1):(Ntip(phy)+Nnode(phy)));

tipstates<-cbind(tipchar01[phy$tip.label],1:Ntip(phy));#colnames(tipstates)<-c("state.t","tipnum")

allstate<-data.frame(rbind(tipstates,parentstates));colnames(allstate)<-c("state","num");rownames(allstate)<-NULL

compar<-c()
for (r in 1:nrow(edges)){
  compar<-c(compar,allstate$state[edges$anc[r]],allstate$state[edges$desc[r]])
}
ancdesc_states<-matrix(compar,ncol=2,byrow = T)
edgestates<-data.frame(cbind(edges,ancdesc_states))
edgestates<-edgestates[order(edgestates$anc),]



#ltt.plot(ctr.complete)
pltt<-ltt.plot.coords(phy)
plot.default(pltt, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r",  type = "S",lwd=3)

#phenogram(phy,tipchar01)
library(dplyr)
anc<-unique(edgestates$anc)
#todo assign correct node ages!
justnode<-seq(from=(Ntip(phy)+1),to=Ntip(phy)+Nnode(phy))
nodedates<-paleotree::dateNodes(phy)[justnode];
nodetimes<-cbind.data.frame(as.numeric(names(nodedates)),nodedates);colnames(nodetimes)<-c("anc","nodetimes")
nodetimes$nodetimes<-nodetimes$nodetimes*-1

edgestates<-left_join(edgestates,nodetimes);colnames(edgestates)<-c("anc","desc","as","ds","nodetimes")
edgestates<-edgestates[order(edgestates$nodetimes),]

cat("Number of LMA->HMA events")
gains<-edgestates$desc[which(edgestates$as==0 &edgestates$ds==1)]
print(length(gains))
cat("Number of HMA->LMA events")
losses<-edgestates$desc[which(edgestates$as==1 &edgestates$ds==0)]
print(length(losses))

#nodeage<-pltt[1:Nnode(phy),1]
#nodestate_age<-as.data.frame(cbind(parentstates,nodeage))
colnames(parentstates)<-c("ASR01","anc")
nodestate_age<-data.frame(parentstates) %>% left_join(.,nodetimes) %>% arrange(nodetimes)

htally<-ltally<-0; hcum<-lcum<-c()
for (r in 1:nrow(nodestate_age)){
  if(nodestate_age$ASR01[r]==0) {
    ltally<-1+ltally
    lcum<-c(lcum,ltally);hcum<-c(hcum,htally)
    }
  if(nodestate_age$ASR01[r]==1 ){
    htally<-1+htally
    hcum<-c(hcum,htally);lcum<-c(lcum,ltally)
    }
}
#tally tips which shifted from MR ancestral state
termEdges<-edgestates[which(edgestates$desc %in% 1:Ntip(phy)),]
l2htips<-termEdges[which(termEdges$as<termEdges$ds),]
h2ltips<-termEdges[which(termEdges$as>termEdges$ds),]
#not sure where to place, in time, so omit 

xy<-as.data.frame(cbind(nodestate_age,lcum,hcum)); xy<-xy[order(xy$nodetimes),]
xyL<-xy[,c("nodetimes","lcum")] 
xyH<-xy[,c("nodetimes","hcum")]


#plot.default(xyL, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r", type = "S",xlim=c(-560,0),ylim=c(0,170),col=lma.col,lwd=3)
lines(xyL, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r", 
        type = "S",col=colors01[1],lwd=3)

lines(xyH, xlab = "Time (MYBP)", ylab = "Number lineages", xaxs = "r", yaxs = "r", 
        type = "S",xlim=c(-560,0),ylim=c(0,170),col=colors01[2],lwd=3)
legend("topleft",legend=c("LMA","HMA"),text.col=colors01,bty="n")
return(list(edgestates,gains, losses))
}
```


```{r tttfunc2,eval=F,results="markdown",fig.height=6}

#todo allow for ambiguous node states
plot_traits_thru_time2<-function(phy,ASR01na,tipchar01,colors01na){

edges<-data.frame(phy$edge);colnames(edges)<-c("anc","desc")

parentstates<-cbind(ASR01na,(Ntip(phy)+1):(Ntip(phy)+Nnode(phy)));#colnames(parentstates)<-c("state.p","nodenum")

tipstates<-cbind(tipchar01[phy$tip.label],1:Ntip(phy));#colnames(tipstates)<-c("state.t","tipnum")

allstate<-data.frame(rbind(tipstates,parentstates));colnames(allstate)<-c("state","num");rownames(allstate)<-NULL

compar<-c()
for (r in 1:nrow(edges)){
  compar<-c(compar,allstate$state[edges$anc[r]],allstate$state[edges$desc[r]])
}
ancdesc_states<-matrix(compar,ncol=2,byrow = T)
edgestates<-data.frame(cbind(edges,ancdesc_states))
edgestates<-edgestates[order(edgestates$anc),]


cat("Number of LMA->HMA events")
gains<-which(edgestates$X1==0 &edgestates$X2==1)
print(length(gains))
cat("Number of HMA->LMA events")
losses<-which(edgestates$X1==1 &edgestates$X2==0)
print(length(losses))



#ltt.plot(ctr.complete)
pltt<-ltt.plot.coords(phy)
plot.default(pltt, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r",  type = "S",lwd=3)

#phenogram(phy,tipchar01)
library(dplyr)
anc<-unique(edgestates$anc)
nodetimes<-data.frame(cbind(anc, pltt[1:Nnode(phy),1]))
edgestates<-left_join(edgestates,nodetimes);colnames(edgestates)<-c("anc","desc","as","ds","nodetimes")
edgestates<-edgestates[order(edgestates$nodetimes),]

nodeage<-pltt[1:Nnode(phy),1]
nodestate_age<-as.data.frame(cbind(parentstates,nodeage))
htally<-ltally<-0; hcum<-lcum<-c()
for (r in 1:nrow(nodestate_age)){
  if(nodestate_age$ASR01[r]==0) {
    ltally<-1+ltally
    lcum<-c(lcum,ltally);hcum<-c(hcum,htally)
    }
  if(nodestate_age$ASR01[r]==1 ){
    htally<-1+htally
    hcum<-c(hcum,htally);lcum<-c(lcum,ltally)
    }
}
#tally tips which shifted from MR ancestral state
termEdges<-edgestates[which(edgestates$desc %in% 1:Ntip(phy)),]
l2htips<-termEdges[which(termEdges$as<termEdges$ds),]
h2ltips<-termEdges[which(termEdges$as>termEdges$ds),]
#not sure where to place, in time, so omit 

xy<-as.data.frame(cbind(nodestate_age,lcum,hcum)); xy<-xy[order(xy$nodeage),]
xyL<-xy[,c("nodeage","lcum")] 
xyH<-xy[,c("nodeage","hcum")]


#plot.default(xyL, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r", type = "S",xlim=c(-560,0),ylim=c(0,170),col=lma.col,lwd=3)
lines(xyL, xlab = "Time", ylab = "N lineages", xaxs = "r", yaxs = "r", 
        type = "S",col=colors01[1],lwd=3)

lines(xyH, xlab = "Time (MYBP)", ylab = "Number lineages", xaxs = "r", yaxs = "r", 
        type = "S",xlim=c(-560,0),ylim=c(0,170),col=colors01[2],lwd=3)
legend("topleft",legend=c("LMA","HMA"),text.col=colors01,bty="n")
return(list(edgestates,gains, losses))
}
```


```{r aceTTT,results="markdown"}
#to use Mk2 marg likelihood estimates
nodeMLstate<-unlist(lapply(bestacemodel$lik.anc[,2],function(x) (ifelse(x<0.5,0,1))))
cat("Using ER model (ACE) to infer ASR")

pdf(paste0(outDir,"figASR.pdf"),width=8,height=11,useDingbats = F)

eventsMk<-plot_traits_thru_time(ctr.complete,nodeMLstate,char01.complete,c(lma.col,hma.col))
title("Using ER model (ACE) to infer ASR")


#nodeMLstate2<-unlist(lapply(bestacemodel$lik.anc[,2],function(x) (ifelse(x<0.75,ifelse(x<0.25,0,NA),1))))
#eventsMk2<-plot_traits_thru_time2(ctr.complete,nodeMLstate2,char01.complete,c(lma.col,hma.col,"gray"))

edges<-eventsMk[[1]]
gains<-unlist(eventsMk[2]); 
losses<-unlist(eventsMk[3])
plot(ctr.complete,cex=0.3,main="MK2 marginals from ACE ER model (calibrated tree)"); axisPhylo()
nodelabels(pie=bestacemodel$lik.anc, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=0.5,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=0.5,frame="none")
nodelabels(node=edges$desc[edges$desc %in% gains],pch=1,lty=1,cex=3,col=hma.col)
nodelabels(node=edges$desc[edges$desc %in% losses],pch=1,lty=1,cex=3,col=lma.col)
dev.off()
####
plot(ctr.complete,cex=0.3,main="MK2 marginals from ACE ER model (calibrated tree)"); axisPhylo()
nodelabels(pie=bestacemodel$lik.anc, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
nodegain<-edges$desc[gains][edges$desc[gains]>Ntip(ctr.complete)]
if(length(nodegain)>0){
 # nodelabels(node=nodegain,text=ctr.complete$node.label[nodegain-Ntip(ctr.complete)],col=hma.col)
  nodelabels(node=nodegain,text=signif(bestacemodel$lik.anc[nodegain-Ntip(ctr.complete),2],2),frame="none",col=hma.col,adj=c(0.5,-0.5))
  }
nodeloss<-edges$desc[losses][edges$desc[losses]>Ntip(ctr.complete)]
if(length(nodeloss)>0){
#nodelabels(node=nodeloss,text=ctr.complete$node.label[nodeloss-Ntip(ctr.complete)],col=lma.col,cex=0.5,bg="white")
 nodelabels(node=nodeloss,text=signif(bestacemodel$lik.anc[nodeloss-Ntip(ctr.complete),1],2),frame="none",col=lma.col,adj=c(0.5,-0.5))
  }



plot(ctr.complete,cex=0.3,main="MK2 marginals from ACE ER model (calibrated tree)"); axisPhylo()
#nodelabels(pie=bestacemodel$lik.anc, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
  nodelabels(text=signif(bestacemodel$lik.anc[,2],2),frame="none",col=hma.col,adj=c(0.5,-0.5),cex=0.5)
 nodelabels(text=signif(bestacemodel$lik.anc[,1],2),frame="none",col=lma.col,adj=c(0.5,0.5),cex=0.5)
#dev.off()
```





## Diversitree analysis of effect HMA/LMA state on host diversification rate

#### Markov 2-state ASR 
Estimate marginal likelihoods for ancestral states under simple Markov model using MCMC and compare to those under BiSSE model (and to ACE models above).  

```{r markov_function,results="markdown"}
## Mk2 ancestral state reconstructions, ignoring the shifts in
## diversification rates:
mcmclength<-10000
runMarkov<-function(tree,char,strict){
	#markov2 MLE alone
  lik.m <- make.mk2(tree, char,strict=strict)
  p <- starting.point.bisse(tree)
  #fit.m.u <- find.mle(lik.m, p[5:6],root=ROOT.FLAT, method="subplex")
    fit.m.u <- find.mle(lik.m, p[5:6])

  cat("Markov estimates for LMA->HMA, HMA->LMA rates")
  print(coef(fit.m.u))
  st<- asr.marginal(lik.m, coef(fit.m.u))
  
    #intial mcmc
	prior <- make.prior.exponential(1 / (2 * (p[1] - p[3])))
	samples<- mcmc(lik.m, coef(fit.m.u), 1000, w=0.1, prior=prior, print.every=100)
	w=diff(sapply(samples[2:3],quantile,c(0.05,0.95)))
  	#Run the final MCMC
    samples.m<-mcmc(lik.m,coef(fit.m.u),nsteps=mcmclength,w=w,prior=prior,print.every=1000)
    st.m <- apply(samples.m[2:3], 1, function(x) asr.marginal(lik.m, x)[2,])
  return(list(st=st,st.m=st.m,samples=samples.m))
}
```

Markov on calibrated: 
```{r mk2_ctr}
markov<-runMarkov(ctr.complete,char01.complete,strict=T)
mk2st.c<-markov$st
mk2st.c2<-markov$st.m
plot(ctr.complete,cex=0.3,main="MK2 MLE only, calibrated tree");axisPhylo()
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
nodelabels(pie=t(mk2st.c), piecol=c(lma.col,hma.col), cex=.5)

cat(" 95% credible intervals for MCMC on Markov model")
#Now we can examine the 95% credible intervals of the posterior samples for each parameter. 
 sapply(markov$samples[,2:3],quantile,c(0.025,0.975))
st.m.avg <- rowMeans(mk2st.c2)

plot(ctr.complete,cex=0.3,main="MCMC marginals from Mk2, calibrated tree");axisPhylo()
nodelabels(pie=1-st.m.avg, piecol=c(lma.col,hma.col), cex=.5)
```




#### BiSSE diversification and character evolution models

Fit models:

- Full, 6 parameter: character gain/loss and speciation & extinction rates for each character state
- Reduced, 5 parameter: single speciation rate, regardless of character state
- Reduced, 5 parameter: single extinction rate, regardless of character state

Plot ASRs using marginal likelihoods from best fitting model.

```{r bisse_function,results="markdown"}
library(diversitree)
library(hisse)
#https://lukejharmon.github.io/ilhabela/2015/07/05/BiSSE-and-HiSSE/
#https://www.zoology.ubc.ca/prog/diversitree/doc/diversitree-manual.pdf
#https://wiki.duke.edu/pages/viewpage.action?pageId=131172442
mcmclength<-10000 #replace wtih 10000 and rn overnight TODO

runBisse<-function(tree,char,strict){
  

equation<-make.bisse(tree, char,strict=strict) # fits full model (6 params) using character state distribution
 p <- starting.point.bisse(tree)

#fit <- find.mle(equation, p,method="subplex",root=ROOT.FLAT) #fit of full character model on tree
fit <- find.mle(equation, p) #fit of full character model on tree

#fit$lnLik
#coef(fit)
lik.l <- constrain(equation, lambda1 ~ lambda0) #alt model where speciation rate doesn't change depending on HMA-LMA state
#fit.l <- find.mle(lik.l, p[argnames(lik.l)],method="subplex")
fit.l <- find.mle(lik.l, p[argnames(lik.l)])
#fit.l$lnLik
#round(rbind(full=coef(fit), equal.l=coef(fit.l, TRUE)), 3)
#anova(fit, equal.l=fit.l) #compare full and constrained models
#tests hypothesis that State0 (LMA) associated with rate increase (2x) compared to state1 (HMA)

cat("Full model")
print(coef(fit))
cat("Constrained model--speciation not affected by character state")
print(coef(fit.l))
cat("ANOVA full vs constrained-lambda model")
print(aov<-anova(fit, lambda.equal=fit.l))
ifelse(aov$"Pr(>|Chi|)"[2]<0.05,bestmodel<-fit,bestmodel<-fit.l)
bestmodel<-fit

lik.2 <- constrain(equation, mu1 ~ mu0) #test that extiction rate doesn't change depending on HMA-LMA state
fit.2<- find.mle(lik.2, p[argnames(lik.2)])

cat("Full model")
print(coef(fit))
cat("Constrained model--extinction not affected by character state")
print(coef(fit.2))
cat("ANOVA full vs constrained-mu model")
print(aov<-anova(fit, mu.equal=fit.2))

# BiSSE MLE ASR 
st <- asr.marginal(equation, coef(bestmodel))

#run MCMC samples
prior <- make.prior.exponential(1 / (2 * (p[1] - p[3])))
initial <- mcmc(equation, bestmodel$par, 1000, w=0.1, prior=prior, print.every=100)

#The tuning parameter for the final MCMC should be equal to approximately the width of the middle 90% of the posterior samples for each parameter value from the preliminary MCMC. We can calculate such a tuning parameter this way:
 
     w=diff(sapply(initial[2:7],quantile,c(0.05,0.95)))
  #Run the final MCMC
 
     samples<-mcmc(equation,bestmodel$par,nsteps=mcmclength,w=w,prior=prior,print.every=1000)
#extract ancestral state posteriors
    st.b <- apply(samples[2:7], 1, function(x) asr.marginal(equation, x)[2,])
return(list(samples=samples,st.b=st.b))
}
```


```{r ctr_bisse,results="markdown"}
if(!(file.exists("bisse_ctr.Rds"))){
	ctr.bisse<-runBisse(ctr.complete,char01.complete,strict=T)
	ctr.bisse.samples<-ctr.bisse$samples
	ctr.bisse.st<-ctr.bisse$st.b
	save(ctr.bisse.samples,ctr.bisse.st,file=paste0(outDir,"bisse_ctr.Rds"))
}else{load(paste0(outDir,"bisse_ctr.Rds") ) }
#plot(ctr.bisse.samples$i, ctr.bisse.samples$p, type="l", xlab="generation", ylab="log(L)")
#colMeans(ctr.bisse.samples)[2:ncol(ctr.bisse.samples)]
#Now we can examine the 95% credible intervals of the posterior samples for each parameter. If the intervals do not overlap, then we have posterior Bayesian support for a difference in rates
 cat("95% CI for speciation, extinction and transition rates")
sapply(ctr.bisse.samples[,2:7],quantile,c(0.025,0.975))
 
#The function above will print a table to your R window. The table of 95% credible intervals indicates that the intervals overlap for each state for speciation, extinction, and transition rate. Thus, we do not infer differences in any parameter that depends on the state of female dispersal.

st.avg <- rowMeans(ctr.bisse.st)
plot(ctr.complete,cex=0.3,main="MCMC marginals from BiSSE model (calibrated tree)"); axisPhylo()
nodelabels(pie=1-st.avg, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
```
```{r profilesCtr, fig.height=6}
#plot speciation rates according to char state
par(mar=c(5,5,1,1))
names(ctr.bisse.samples)[2:3]<-c("LMA_l","HMA_l")
profiles.plot(ctr.bisse.samples[c("LMA_l", "HMA_l")], col.line=c(lma.col,hma.col), las=1, xlab="Speciation rate (lamda)", legend="topright")

names(ctr.bisse.samples)[4:5]<-c("LMA_m","HMA_m")
profiles.plot(ctr.bisse.samples[c("LMA_m", "HMA_m")], col.line=c(lma.col,hma.col), las=1, xlab="Extinction rate (mu)", legend="topright")

#Net diversification plot

net.div<-ctr.bisse.samples[,c("LMA_l","HMA_l")]-
    ctr.bisse.samples[,c("LMA_m","HMA_m")]
colnames(net.div)<-paste("lambda-mu(",c("LMA","HMA"),")",sep="")
profiles.plot(net.div, 
    xlab="Net diversification rate", ylab="Probability density",
    legend.pos="topleft",col.line=c(lma.col,hma.col),
    lty=1)
sapply(net.div[,1:2],quantile,c(0.025,0.975))

```



##Lineages-thru-time for each state
Using mean marginal likelihood states from BiSSE full `r mcmclength` MCMC samples.

```{r lttbisse,results="markdown"}

#to use Mk2 marg likelihood estimates
nodeMLstate<-unlist(lapply(st.m.avg,function(x) (ifelse(x<0.5,0,1))))
cat("Using Mk2 model (MCMC) to infer ASR")
eventsMk<-plot_traits_thru_time(ctr.complete,nodeMLstate,char01.complete,c(lma.col,hma.col))
title("Using Mk2 model (MCMC) to infer ASR")

#to use BiSSE MCMC marginals
cat("Using BiSSE model (MCMC) to infer ASR")
nodeMLstate<-unlist(lapply(st.avg,function(x) (ifelse(x<0.5,0,1))))
events<-plot_traits_thru_time(ctr.complete,nodeMLstate,char01.complete,c(lma.col,hma.col))
title("Using BiSSE model (MCMC) to infer ASR")
```
```{r annotEvents}
edges<-events[[1]]
gains<-unlist(events[2])
losses<-unlist(events[3])
plot(ctr.complete,cex=0.3,main="MCMC marginals from BiSSE model (calibrated tree)"); axisPhylo()
nodelabels(pie=1-st.avg, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
nodelabels(node=edges$desc[gains],pch=1,lty=1,cex=3,col=hma.col)
nodelabels(node=edges$desc[losses],pch=1,lty=1,cex=3,col=lma.col)


plot(ctr.complete,cex=0.3,main="MCMC marginals from BiSSE model (calibrated tree)"); axisPhylo()
nodelabels(pie=1-st.avg, piecol=c(lma.col,hma.col), cex=.5)
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==0]),col=lma.col,pch=16,cex=1,frame="none")
tiplabels(tip=which(ctr.complete$tip.label %in% names(char01.complete)[char01.complete==1]),col=hma.col,pch=16,cex=1,frame="none")
nodegain<-edges$desc[gains][edges$desc[gains]>Ntip(ctr.complete)]
if(length(nodegain)>0){
  nodelabels(node=nodegain,text=ctr.complete$node.label[nodegain-Ntip(ctr.complete)],col=hma.col)
}
nodeloss<-edges$desc[losses][edges$desc[losses]>Ntip(ctr.complete)]
if(length(nodeloss)>0){

nodelabels(node=nodeloss,text=ctr.complete$node.label[nodeloss-Ntip(ctr.complete)],col=lma.col,cex=0.5,bg="white")
}

```

