---
title: "Subtraction of seawater ASVs from sponge microbial counts"
output: 
  html_document:
  toc: true
toc_float: true
collapsed: false
toc_depth: 4
df_print: paged
author: "Sabrina Pankey"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE,fig.width = 8, results="markdown")
options(scipen = 999)
```


```{r load, echo=FALSE, include=FALSE} 
mainDir<-"~/Box Sync/MPL/Lesser DOB sponges/coevolution_analyses/pipeline4/"
setwd(mainDir)
outDir<-"06_swadj_results/"
ifelse(!dir.exists(file.path(mainDir, outDir)), dir.create(file.path(mainDir, outDir)), FALSE)
library(phyloseq,warn.conflicts = FALSE)
library(reshape2)
library(ape)
#inputs: 
# - either: ps4.rds or ps.adj.RDS
# - species_charsRF.Rds

#outputs: 
# ps.adj.RDS (only if starting with ps4)    -seawater-normalised counts for each sample
# ps.adj.hostspecies.counts.RDS             -seawater-normalised counts for merged by species (not rarefied)
# ps.adj.hostspecies.ra.RDS                     -seawater-normalised relative abundance for merged by species (after count rarefaction)
options(getClass.msg=FALSE)
```
```{r subtract,echo=FALSE, include=FALSE}
if(file.exists(paste0(outDir,"ps.adj.RDS"))){load(paste0(outDir,"ps.adj.RDS")) #loads previously-made ps.adj with seawater-corrected counts
  
}else{
  load("01_results/ps4.rds")
  
  #sort sample_data(ps4) to mirror sponge phylogeny
  df<-data.frame(sample_data(ps4)[order(as.numeric(sample_data(ps4)$phyloTreeRank))])
  df<-df[df$X!="",];df<-df[df$X!="#N/A",]
  df<-df[order(as.numeric(as.character(df$X))),]
  Species.order<-c(as.character(unique(df$Species_consensus)),"seawater","sediment")
  Genera.order<-unique(df$phyloGenus)
  Family.order<-unique(df$phyloFamily)
  Order.order<-unique(df$phyloOrder)
  
  #seawater taxa--use abundance to normalize abundance in sponges?
  # for each ASV in seawater- calc relative abundance
  # calc expected number of counts for this ASV in each sample due to seawater contribution
  # subtract: observed - expected, if negative, set to 0.
  ps4.sw<-prune_samples(sample_data(ps4)$Species_consensus=="seawater",ps4)
  pssw.m<-merge_samples(ps4.sw,"locale",sum)
  pssw.m<-prune_taxa(taxa_names(pssw.m)[which(taxa_sums(pssw.m)>0)],pssw.m)
  pssw.m.ra <- transform_sample_counts(pssw.m, function(OTU) OTU/sum(OTU)) #convert to relative abundance
  head(otu_table(pssw.m.ra)[,1:10])
  
  subtractSWcounts<-function(loc,ps.seawater_by_loc.ra,psobject)
  {
    loc.sw.ra<-data.frame(otu_table(ps.seawater_by_loc.ra)[loc,]); loc.sw.ra<-loc.sw.ra[which(loc.sw.ra>0)]
    loc.samp<-prune_samples(sample_data(psobject)$locale==loc,psobject)
    loc.samp<-prune_samples(sample_data(loc.samp)$Species_consensus!="seawater",loc.samp)
    loc.samp<-prune_taxa(taxa_names(loc.samp)[which(taxa_sums(loc.samp)>0)],loc.samp)
    locCounts<-data.frame(otu_table(loc.samp)); dim(locCounts)
    locCountsSW<-locCounts[colnames(loc.sw.ra),]; dim(locCountsSW)
    locCounts_rest<-locCounts[!(rownames(locCounts) %in% rownames(locCountsSW)),];dim(locCounts_rest)
    adjusted<-apply(t(locCountsSW), 1, FUN= function(x) x-(colSums(locCounts)*loc.sw.ra) )
    adjusted<-do.call(rbind,adjusted)
    adjusted[adjusted<0]<-0
    adjusted<-t(adjusted); dim(adjusted)
    adjCounts <-rbind(locCounts_rest,adjusted); dim(adjCounts)
    return(adjCounts)
  }
  BZadjCounts<-subtractSWcounts("BZ",pssw.m.ra,ps4); BZadjCounts<-t(BZadjCounts); samp<-rownames(BZadjCounts); BZadjCounts<-cbind.data.frame(samp,BZadjCounts,stringsAsFactors=F)
  CUadjCounts<-subtractSWcounts("CU",pssw.m.ra,ps4); CUadjCounts<-t(CUadjCounts); samp<-rownames(CUadjCounts); CUadjCounts<-cbind.data.frame(samp,CUadjCounts,stringsAsFactors=F)
  KYadjCounts<-subtractSWcounts("KY",pssw.m.ra,ps4); KYadjCounts<-t(KYadjCounts); samp<-rownames(KYadjCounts); KYadjCounts<-cbind.data.frame(samp,KYadjCounts,stringsAsFactors=F)
  SXadjCounts<-subtractSWcounts("SX",pssw.m.ra,ps4); SXadjCounts<-t(SXadjCounts); samp<-rownames(SXadjCounts); SXadjCounts<-cbind.data.frame(samp,SXadjCounts,stringsAsFactors=F)
  
  #merge site-ajusted counts in to one table:
  m1<-merge(BZadjCounts,CUadjCounts,all=T); dim(m1); dim(BZadjCounts); dim(CUadjCounts)
  m1<-merge(m1,KYadjCounts,all=T); dim(m1); dim(KYadjCounts)
  m1<-merge(m1,SXadjCounts,all=T); dim(m1); dim(SXadjCounts)
  
  swdf<-t(data.frame(otu_table(ps4.sw))); dim(swdf);samp<-rownames(swdf); swdf<-cbind.data.frame(samp,swdf,stringsAsFactors=F)
  m2<-merge(m1,swdf,all=T); dim(m2)
  m2$samp<-gsub("\\.","-",m2$samp)
  
  rownames(m2)<-m2$samp; m3<-m2[,-1]
  m3[is.na(m3)]<-0
  
  
  rownames(m3) %in% sample_names(ps4)
  colnames(m3) %in% taxa_names(ps4)
  taxa_names(ps4)[!(taxa_names(ps4) %in% colnames(m3))]; colnames(m3)[!(colnames(m3) %in% taxa_names(ps4) )]
  #create phyloseq object with sw-adjusted sponge counts
  
  ps.adj<-phyloseq(otu_table(as.matrix(m3),taxa_are_rows=F),sample_data(ps4),tax_table(ps4))
  sample_sums(ps.adj)
  hist(sample_sums(ps.adj), breaks=100)
  
  save(ps.adj,file=paste0(outDir,"ps.adj.RDS"))
}

if(!file.exists("ps.adj.hostspecies.counts.RDS")){

#merge samples by species
#sample_sums(ps.adj)
ps.adj<-prune_samples(sample_data(ps.adj)$Species_consensus!="",ps.adj) 

ps.adj.hostspecies<-merge_samples(ps.adj,"Species_consensus") 
sample_data(ps.adj.hostspecies)$Species_consensus<-as.factor(sample_names(ps.adj.hostspecies)) #repair factor levels

set.seed(1234)
#look at species with lowest counts
#sort(sample_sums(ps.adj.hostspecies), F)[1:10]
#sample_data(ps.adj) %>% group_by(Species_consensus) %>% count()  %>% print(n = Inf)
ps.adj.hostspecies<-rarefy_even_depth(ps.adj.hostspecies)#repeat normalization because PhyloSeq sums reads across merged samples
ps.adj.hostspecies<-prune_taxa(taxa_sums(ps.adj.hostspecies)>0,ps.adj.hostspecies)

#sample_data(ps.adj.hostspecies)$phyloFamily <- levels(sample_data(ps.adj)$phyloFamily)[get_variable(ps.adj.hostspecies,"phyloFamily")]
#sample_data(ps.adj.hostspecies)$phyloOrder <- levels(sample_data(ps.adj)$phyloOrder)[get_variable(ps.adj.hostspecies,"phyloOrder")]
#sample_data(ps.adj.hostspecies)$phyloGenus <- levels(sample_data(ps.adj)$phyloGenus)[get_variable(ps.adj.hostspecies,"phyloGenus")]
#sample_data(ps.adj.hostspecies)$phylosubClass <- levels(sample_data(ps.adj)$phylosubClass)[get_variable(ps.adj.hostspecies,"phylosubClass")]
#sample_data(ps.adj.hostspecies)$phyloClass <- levels(sample_data(ps.adj)$phyloClass)[get_variable(ps.adj.hostspecies,"phyloClass")]

#replace sample_data with character and phylo info
load("02_tree_results/spongespeciesTimeTree.OGless.Rds") #timetr #treemeta.u
#species phylo-ordered

timetr.dob<-keep.tip(timetr,tip=timetr$tip.label [timetr$tip.label %in% sample_names(ps.adj.hostspecies)])
p<-ggtree::ggtree(timetr.dob,right=T)
Species.order<-c(ggtree::get_taxa_name(p),"seawater","sediment")

ps.adj.hostspecies<-prune_samples(Species.order,ps.adj.hostspecies)
sample_data(ps.adj.hostspecies)$Species_consensus<- factor(sample_data(ps.adj.hostspecies)$Species_consensus, levels = Species.order)
dat1<-data.frame(sample_data(ps.adj.hostspecies))

#head(dat1); tail(dat1)

load("03_char_results/species_charsRF.Rds") #use char$consensusRF_phyla
library(dplyr)
dat2<-left_join(dat1,char, by=c("Species_consensus"))

dat2<-dat2 %>% select(Species_consensus,consensusmod,consensusRF_asv,consensusRF_class,consensusRF_phyla)
dat2$RFprediction<-dat2$consensusRF_phyla


dat3<-left_join(dat2,treemeta.u); rownames(dat3)<-dat3$Species_consensus
sample_data(ps.adj.hostspecies)<-dat3

ps.adj.hostspecies<-prune_taxa(taxa_sums(ps.adj.hostspecies)>0,ps.adj.hostspecies)


save(ps.adj.hostspecies,file=paste0(outDir,"ps.adj.hostspecies.counts.RDS"))


ps.adj.hostspecies.ra<-transform_sample_counts(ps.adj.hostspecies, function(x) x / sum(x))
sample_data(ps.adj.hostspecies.ra)$Species_consensus<-factor(sample_names(ps.adj.hostspecies.ra),levels=Species.order) #repair factor levels
sample_sums(ps.adj.hostspecies.ra)
ntaxa(ps.adj.hostspecies.ra)
nsamples(ps.adj.hostspecies.ra)

save(ps.adj.hostspecies.ra,file=paste0(outDir,"ps.adj.hostspecies.ra.RDS"))


#comparison original non-adjusted counts:


}else{ load(paste0(outDir,"ps.adj.hostspecies.counts.RDS"))} #ps.adj.hostspecies
```

### Composition of all species, before and after SW subtraction
```{r comparisons1,echo=FALSE, include=FALSE}
  load("01_results/psRm_bySpecies.rds") #psRm
  load("01_results/microColors.rds") #commonphyla_colors,commonclass_colors
  load("02_tree_results/spongespeciesTimeTree.OGless.Rds") #timetr
 library(ape)
  psRm<-prune_samples(sample_names(psRm) %in% timetr$tip.label,psRm)
  psRm.ra<-transform_sample_counts(psRm, function(x) x / sum(x))
 ps.adj.hostspecies<-prune_samples(sample_names(ps.adj.hostspecies) %in% timetr$tip.label,ps.adj.hostspecies)
 ps.adj.hostspecies<-prune_taxa(taxa_sums(ps.adj.hostspecies)>0, ps.adj.hostspecies)
   
  
  
#determine what top phyla are shared
 top <- names(sort(taxa_sums(ps.adj.hostspecies), decreasing=TRUE))[1:3000]
  sum(taxa_sums(ps.adj.hostspecies)[top])/sum(taxa_sums(ps.adj.hostspecies))
  topnames.adj<-as(unique(tax_table(ps.adj.hostspecies)[top,"KP"]),"vector")
  orignames<-names(commonphyla_colors)[!(names(commonphyla_colors) %in% topnames.adj)]
  adjnames<-topnames.adj [!(topnames.adj  %in% names(commonphyla_colors) )]
```
FOr both origial and SW-corrected data, top 3000 most abundant ASVs include members of:
`r intersect(topnames.adj,names(commonphyla_colors))`

Original counts also include members of `r orignames`.

SW-corrected counts find ASVs from `r adjnames` to be more abundant:


```{r comparebars}
topglom_order_ra<-function(ps,tree,topnames,glomRank) {
  # tree<-timetr;topnames<-names(commonphyla_colors);glomRank<-"Phylum";ps<-ps.adj.hostspecies
  
  top <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:3000]
  ps.top <- prune_taxa(top, ps)
  
 #glom
  glom<-tax_glom(ps.top,taxrank=glomRank)
  if(glomRank=="Phylum"){tax_table(glom)[,"KP"]<-paste0(tax_table(glom)[,"Kingdom"],"_",tax_table(glom)[,"Phylum"])}
  if(glomRank=="Class"){tax_table(glom)[,"KPC"]<-paste0(tax_table(glom)[,"Kingdom"],"_",tax_table(glom)[,"Phylum"],"_",tax_table(glom)[,"Class"])}
  
  psra<-transform_sample_counts(glom, function(x) x / sum(x))
  sample_data(psra)$Species_consensus<-as.factor(sample_names(psra)) #repair factor levels
  
 
  #species phylo-ordered
  timetr.dob<-keep.tip(tree,tip=tree$tip.label [tree$tip.label %in% sample_names(psra)])
  p<-ggtree::ggtree(timetr.dob,right=T)
  sample_data(psra)$Species_consensus<- factor(sample_data(psra)$Species_consensus, levels = ggtree::get_taxa_name(p))
  return(psra)
}
```

```{r bars}
library(colorspace)
glomA<-topglom_order_ra(ps.adj.hostspecies,timetr,names(commonphyla_colors),"Phylum")
glomO<-topglom_order_ra(psRm,timetr,names(commonphyla_colors),"Phylum")
library(ggplot2)
   # PHYLUM: due to size, plot bars and legend separately
  pbarO<-plot_bar(glomO, x="Species_consensus", fill="KP")  + scale_fill_manual(values = commonphyla_colors) + guides(fill=guide_legend(ncol=2))
  pbarO  + ggtitle("Relative abundance of 3000 most common ASVs (by PHYLUM on original counts)") + theme(legend.position="none",axis.text=element_text(size=6))
   # extract Legend to plot below
  legend <- cowplot::get_legend(pbarO); cowplot::ggdraw(legend)
 
  cols32<-c(rainbow_hcl(8, l=30,start = 10, end = 300),
          rainbow_hcl(8, l=50,start = 10, end = 300), 
          rainbow_hcl(8, l=70,start = 10, end = 300), 
          rainbow_hcl(8, l=90,start = 10, end = 300))
  adjcols<-c(cols32[!(cols32 %in% commonphyla_colors)][1:length(adjnames)], commonphyla_colors[names(commonphyla_colors) %in% topnames.adj ]);
  names(adjcols)[1:length(adjnames)]<-adjnames
  pbarA<-plot_bar(glomA, x="Species_consensus", fill="KP")  + scale_fill_manual(values = adjcols) + guides(fill=guide_legend(ncol=2))
  pbarA  + ggtitle("Relative abundance of 3000 most common ASVs (by PHYLUM on SW-corrected counts)") + theme(legend.position="none",axis.text=element_text(size=6))
  
  # extract Legend to plot below

   legend <- cowplot::get_legend(pbarA); cowplot::ggdraw(legend)
  

  psaH3<-prune_samples(sample_data(glomA)$RFprediction=="HMA",glomA)
  psaL3<-prune_samples(sample_data(glomA)$RFprediction=="LMA",glomA)
  
  psoH3<-prune_samples(sample_names(psaH3),glomO)
  psoL3<-prune_samples(sample_names(psaL3),glomO)
  
```

### Composition of HMA and LMA species, SW-corrected
```{r barplots}
# pdf(paste0(outDir,"fig2_barplots.pdf"))
  p.ha<-plot_bar(psaH3, x="Species_consensus", fill="KP") +    scale_fill_manual(values = adjcols)
  p.ha  + theme(legend.position="none",axis.text=element_text(size=6)) 
  p.la<-plot_bar(psaL3, x="Species_consensus", fill="KP") +    scale_fill_manual(values = adjcols)
  p.la  + theme(legend.position="none",axis.text=element_text(size=6)) 
  
   # extract Legend to plot below
  legend <- cowplot::get_legend(p.la); cowplot::ggdraw(legend)
 # dev.off()
```


```{r anosim}
library(vegan)
#test whether distances within group less than differences between groups
ano<-vegan::anosim(otu_table(ps.adj.hostspecies),sample_data(ps.adj.hostspecies)$RFprediction)

ano
#test whether distance differs btween groups
dist<-phyloseq::distance(ps.adj.hostspecies,method="bray")
samp<-as(sample_data(ps.adj.hostspecies),"data.frame")
ado<-adonis(dist ~ RFprediction,data=samp,permutations = 9999)
ado
# Homogeneity of dispersion test
beta <- betadisper(dist, samp$RFprediction)
bperm<-permutest(beta)
TukeyHSD(beta)

```
ANOSIM finds LMA and HMA are compositionally distinct (ASV abundances):
R= `r ano$statistic`, p= `r ano$signif`. 
Simlilarly, PERMANOVA finds signficantly different composititon:F=`r ado$aov.tab$F.Model[1]`,R2=`r ado$aov.tab$R2[1]`,p=`r ado$aov.tab$ "Pr(>F)"[1]`.

And testing homogeneity of dispersion finds that LMA sponges are signifcantly more variable in composition than HMA: F=`r bperm$statistic`,p=`r bperm$tab$ "Pr(>F)"[1]`.

### Effect of SW correction on abundances of common phyla  
```{r comparisons, results="markdown"}

hma.col<-rainbow_hcl(12)[2]; lma.col<-rainbow_hcl(12)[7]

  orig.ra<-apply(otu_table(glomO),2,mean)
  names(orig.ra)<-tax_table(glomO)[,"KP"]
  adj.ra<-apply(otu_table(glomA),2,mean)
  names(adj.ra)<-tax_table(glomA)[,"KP"]
  adj.ra<-adj.ra[names(orig.ra)]
  comparison<-cbind.data.frame(orig.ra,adj.ra)
  comparison$KP<-names(orig.ra)
  
  
  orig.ra.l<-apply(otu_table(psoL3),2,mean)
names(orig.ra.l)<-tax_table(psoL3)[,"KP"]
orig.ra.h<-apply(otu_table(psoH3),2,mean)
names(orig.ra.h)<-tax_table(psoH3)[,"KP"]

adj.ra.h<-apply(otu_table(psaH3),2,mean)
names(adj.ra.h)<-tax_table(psaH3)[,"KP"]
adj.ra.l<-apply(otu_table(psaL3),2,mean)
names(adj.ra.l)<-tax_table(psaL3)[,"KP"]

ol<-data.frame(orig.ra.l)
al<-data.frame(adj.ra.l)
oh<-data.frame(orig.ra.h)
ah<-data.frame(adj.ra.h)
comparison.l<-merge(ol,al,all.x = T,all.y = T,by=0)
colnames(comparison.l)[1]<-"KP"
comparison.h<-merge(oh,ah,all.x = T,all.y = T,by=0)
colnames(comparison.h)[1]<-"KP"
comparison.l[is.na(comparison.l)]<-comparison.h[is.na(comparison.h)]<-0
library(tidyr)
#melts multiple columns to single column of values to plot with accompany column 'process' pulled from column names
plotsideBars<-function(dat,cols){
  gathered<-gather(dat,process,value, -KP)
  #gathered$KP<-factor(gathered$KP,levels=unique(dat$KP))
  gathered$KP<-factor(gathered$KP)
  
  ggplot(gathered, aes(x=KP, y=value)) +   
    geom_bar(aes(fill = process), position = "dodge", stat="identity") +
    theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +  scale_fill_manual(values=cols)
  
}

p0<-plotsideBars(comparison,c("gray","black"))
p0 + ggtitle("before v after SW correction")
wilcox.test(comparison$orig.ra,comparison$adj.ra,paired = T)

p1<-plotsideBars(comparison.l,c("gray","black"))
p1 + ggtitle("LMA species only, before v after SW correction")
wilcox.test(comparison.l$orig.ra,comparison.l$adj.ra,paired = T)

p2<-plotsideBars(comparison.h,c("gray","black"))
p2 +ggtitle("HMA species only, before v after SW correction")
wilcox.test(comparison.h$orig.ra,comparison.h$adj.ra,paired = T)

comparison.orig<-cbind.data.frame(orig.ra.l,orig.ra.h)
comparison.orig$KP<-names(orig.ra.l)
p3<-plotsideBars(comparison.orig,c(hma.col,lma.col))
p3 +ggtitle("HMA v LMA, before SW correction")
#wilcox.test(comparison.orig$orig.ra.l,comparison.orig$orig.ra.h,paired = T)

comparison.adj<-cbind.data.frame(adj.ra.l,adj.ra.h)
comparison.adj$KP<-names(adj.ra.l)
p4<-plotsideBars(comparison.adj,c(hma.col,lma.col))
p4 +ggtitle("HMA v LMA, after SW correction")
#wilcox.test(comparison.adj$adj.ra.l,comparison.adj$adj.ra.h,paired = T)

```

