---
title: "Microbial symbiont diversification"
output: 
  html_document:
  toc: true
toc_float: true
collapsed: false
toc_depth: 4
df_print: paged
author: "Sabrina Pankey"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE, message=FALSE,fig.width = 8, results="markdown")
options(scipen = 999)
```

```{r setup2}
mainDir<-"~/Box Sync/MPL/Lesser DOB sponges/coevolution_analyses/pipeline4/"
setwd(mainDir)
outDir<-"07_divresults/"
dir.exists(file.path(mainDir, outDir))
library(phyloseq,warn.conflicts = FALSE)
library(ape,warn.conflicts = FALSE)
library(colorspace)
hma.col<-rainbow_hcl(12)[2]; lma.col<-rainbow_hcl(12)[7]
library(vegan)

library(ggplot2)
library(ggtree,warn.conflicts = FALSE)
options(getClass.msg=FALSE)


#input: 
# sw-corrected relative abundance phyloseq
# RFimportance_phyla.Rds"  #loads 'imp.phy' with importance scores for how predictive each phylum is for HMA or LMA classification
# RFimportance_class.Rds"
load(file="06_swadj_results/ps.adj.hostspecies.ra.RDS")  # ps.adj.hostspecies.ra
psa<-ps.adj.hostspecies.ra
psa<-prune_taxa(taxa_sums(psa)>0,psa)
load("02_tree_results/spongespeciesTimeTree.OGless.Rds") #timetr
#species phylo-ordered
timetr.dob<-keep.tip(timetr,tip=timetr$tip.label [timetr$tip.label %in% sample_names(psa)])
p<-ggtree::ggtree(timetr.dob,right=T)
Species.order<-c(ggtree::get_taxa_name(p),"seawater","sediment")


sample_data(psa)$phyloOrder %in% names(cladecols)
cladecols<-cladecols[names(cladecols) %in% sample_data(psa)$phyloOrder]

treemeta.u<-treemeta.u[treemeta.u$Species_consensus %in% timetr.dob$tip.label,]
treemeta.u$species_color<-cladecols[match(treemeta.u$phyloOrder,names(cladecols)) ] 
  
sample_data(psa)$phyloOrder[sample_data(psa)$Species_consensus=="seawater"]<-"seawater"
sample_data(psa)$phyloOrder[sample_data(psa)$Species_consensus=="sediment"]<-"sediment"

sample_data(psa)$phyloOrder<-factor(sample_data(psa)$phyloOrder,levels=c(names(cladecols),"seawater","sediment"))
env_cladecols<-c(cladecols,"gray","black"); names(env_cladecols)<-c(names(cladecols),"seawater","sediment")
```

```{r setupdata_forrichness}
load("06_swadj_results/ps.adj.RDS")
load("03_char_results/species_charsRF.Rds") #use char$consensusRF_phyla
#species phylo-ordered
tmp<-otu_table(ps.adj)
#otu_table(ps.adj)<-otu_table(round(tmp),taxa_are_rows = F)

keep<-sample_names(ps.adj)[sample_data(ps.adj)$Species_consensus %in% Species.order]
ps.adj<-prune_samples(keep,ps.adj)
sample_data(ps.adj)$Species_consensus<- factor(sample_data(ps.adj)$Species_consensus, levels = Species.order)
dat1<-data.frame(sample_data(ps.adj))
rnams<-rownames(dat1)
library(dplyr)
dat2<-left_join(dat1,char, by=c("Species_consensus"))
dat2<-dat2 %>% select(Species_consensus,consensusmod,consensusRF_asv,consensusRF_class,consensusRF_phyla)
dat2$RFprediction<-dat2$consensusRF_phyla
dat3<-left_join(dat2,treemeta.u); rownames(dat3)<-rnams
sample_data(ps.adj)<-dat3
set.seed(1234)
ps.adjR<-rarefy_even_depth(ps.adj,sample.size = 8000)
ps.adjR<-prune_taxa(taxa_sums(ps.adjR)>0,ps.adjR)
ps.adjR<-prune_samples(sample_sums(ps.adjR)>0,ps.adjR)


#use ps.adjR to calc per-sample richness, then average
```
##Calculating richness and evenness
For sponges, calculate alpha diversities of symbionts for each species
For microbes, calcuate alpha diversities of host habitats for each ASV

- Total N: absolute richness
- Shannon-Weiner H' (abundance-informed richness): increases with both evenness and richness, so difficult to compare comm that differ in richness
- Simpson's D(evenness/dominance): less sensitive to richness (as D increases to 1, diversity decreases (ie monoculture); 1-D: higher means more diverse)
- Pielou evenness (ratio of diversity index to maximum possible diversity)
```{r calcSHD}

### H1: Specialized ASVs more often found in HMA sponges.

# 1. calc 'specialist index' for each ASV, 
# -visual:map these values, and host species, to bacterial trees...is this a phylogenetically conserved 'trait'? 
# ie, what is the phylogenetic signal for being specialist/generalist?
# - 
  # Are any types of microbes more specialized overall?
# -compare mean specialization between HMA and LMAs, consider indep constrasts to account for phylogeny.
  # are ASVs in HMA sponges characterized by lower mean richness, H,D,evenness than the ASVs of LMAs?

#get averages to plot for tree
spec <- factor(sample_data(ps.adjR)$Species_consensus)
  #sponge sample richness
  #number ASVs per SAMPLE, then average for species
  S.samp <- apply(otu_table(ps.adjR)>0,1,sum) # S = total number of ASVs with non-zero abundance in a xMA sponge SAMPLE
  S.sponge <- sapply(levels(spec),function(i){mean(S.samp[spec==i])}) #species average richness
  shannons.samp<-apply(otu_table(ps.adjR),1,vegan::diversity)
  shannons.sponge<- sapply(levels(spec),function(i){ mean(shannons.samp[spec==i])}) #species average
  plot(S.sponge,shannons.sponge) ;title("Sponge: ASV Richness vs Shannon's H index")
  simpsons.samp<-apply(otu_table(ps.adjR),1, FUN=function(x) vegan::diversity(x,index="simpson")) #returns 1-D
  simpsons.sponge<-sapply(levels(spec),function(i){ mean(simpsons.samp[spec==i])}) #species average
  plot(S.sponge,simpsons.sponge); title("Sponge: ASV Richness vs Simpsons's D index")
  pielousH.samp<-shannons.samp/log(S.samp)
  pielousH.sponge<- sapply(levels(spec),function(i){ mean(pielousH.samp[spec==i])})
  plot(S.sponge,pielousH.sponge); title("Sponge: ASV Richness vs Pielou's evenness using H")
  pielousD.samp<-simpsons.samp/log(S.samp)
    pielousD.sponge<-sapply(levels(spec),function(i){ mean(pielousD.samp[spec==i])})
  plot(S.sponge,pielousD.sponge); title("Sponge: ASV Richness vs Pielou's evenness using D")
  plot(pielousH.sponge,pielousD.sponge); title("Sponge: Pielou's evenness using H vs using D")

  
  
  #ASV host habitat richness
  #number host SPECIES per asv (use species merged-counts)
  S.ASV <- apply(otu_table(psa)>0,2,sum) # S = total number of host species with non-zero occurence of an ASV
  shannons.ASV<-apply(otu_table(psa),2,vegan::diversity)
  plot(S.ASV,shannons.ASV); title("ASV: Host Richness vs Shannon's H index")
  simpsons.ASV<-apply(otu_table(psa),2, FUN=function(x) vegan::diversity(x,index="simpson")) #returns 1-D
  plot(S.ASV,simpsons.ASV); title("ASV: Host Richness vs Simpsons's D index")
  pielousH.ASVs<-shannons.ASV/log(S.ASV+1)  #add 1 to prevent NaN when S.ASV is only 1 host (log(1)=0)
  plot(S.ASV,pielousH.ASVs); title("ASV: Host Richness vs Pielou's evenness using H")
  pielousD.ASVs<-simpsons.ASV/log(S.ASV+1)
  plot(pielousH.ASVs,pielousD.ASVs)
  title("ASV: Pielou's evenness using H vs using D")

  plot(S.ASV,pielousD.ASVs)
  title("ASV: Host Richness vs Pielou's evenness using D")

#Simpsons values less weighted by high richness values.

  
  
#calc H' shannon-weiner(abundance-informed richness): increases with both evenness and richness, so difficult to compare comm that differ in richness
#less sensitive to higher richness

#calc D Simpson(evenness/dominance): less sensitive to richness (as D increases to 1, diversity decreases (ie monoculture); 1-D: higher means more diverse)
#not well related
#very non linear

#calc Pielou evenness (ratio of diversity index to maximum possible diversity)




#skip Pielous and Simpsons
```

##Phylogenetic signal of alpha diversities in microbe trees
Omitting Simpsons and Pielous since they highly upweight low-abundances.
Maps metrics to ASV tips and calcuates Pagel's $\lambda for each metric.
Maps hosts occupied (and abundance therein) to each ASV tip. Plots saved to `07_divresults` folder.
```{r}
#map values onto ASV trees:
#tree=tr;ps=psa; richness=S.ASV; shannons=shannons.ASV;  treetitle=title
calc_div_signal<-function(tree, ps, richness,shannons,simpsons, pielous,treetitle){
  tree<-keep.tip(tree,tip=tree$tip.label[tree$tip.label %in% taxa_names(ps)])
 
   if(Ntip(tree)>2){
     ps<-prune_taxa(tree$tip.label,ps)
     ps<-merge_phyloseq(sample_data(ps),tax_table(ps),otu_table(ps),tree)
     
  N<-richness[names(richness) %in% tree$tip.label]
  H<-shannons[names(shannons) %in% tree$tip.label]

  div.df<-cbind.data.frame(names(N),N,H)
  mean_divs<-apply(div.df[,2:3],2,FUN= function(x) mean(x,na.rm=T))

    if(sum(sample_sums(ps)[!(sample_names(ps) %in% c("seawater","sediment"))])>0){
  sample_data(ps)$Species_consensus<-factor(sample_names(ps),levels=Species.order)
 
  #subset phyloORder colors needed
  coltmp<-cladecols[names(cladecols) %in% sample_data(ps)$phyloOrder]
  #plot host occurence across ASV tips, colored by phyloegenetic clade
  pt<-plot_tree(ps,size="abundance",color="phyloOrder",label.tips = "taxa_names") + scale_color_manual(values=c(coltmp,"gray","black"))
  pt.asv<-pt + theme(legend.position="none") +  ggtitle(treetitle) 
  
  legend <- cowplot::get_legend(pt ); leg<-cowplot::ggdraw(legend)
  pt.heat<-plot_heatmap(ps,taxa.order=tree$tip.label,sample.order=Species.order)
  
  #plot sponge type(high or low MA) across ASV tips
  pt<-plot_tree(ps,color="RFprediction",label.tips = "taxa_names") + scale_color_manual(values=c(hma.col,lma.col,"gray","black"))
  pt.asvRF<-pt + ggtitle(treetitle) 
  
  #set up ASV tree
  pt<-ggtree(phy_tree(ps)) + geom_tiplab(size=1,align=F)
  
  
  #plot observed habitat richness across ASV tips and calc signal (lambda) 
  p4d.N<-phylobase::phylo4d(phy_tree(ps), tip.data=div.df$N)
  lambda.N<-as.character(round(phylosignal::phyloSignal(p4d.N,methods="Lambda")$stat,3))
  
  pt.N<-pt %<+% div.df + geom_tippoint(aes(color=N),size=2) + theme_tree2() + scale_color_continuous(high='#D55E00',low='#0072B2')  +
    theme(legend.position="right") +ggtitle(paste0(treetitle," (Observed richness of host habitat')")) 
   pt.N<-pt.N + annotate("text",x=1,y=5,label=paste0("Lambda: ",lambda.N)) 
   
  #plot shannon evenness  and its phylo signal (lambda) across ASV tips
  p4d.H<-phylobase::phylo4d(phy_tree(ps), tip.data=div.df$H)
  lambda.H<-as.character(round(phylosignal::phyloSignal(p4d.H,methods="Lambda")$stat,3))
  pt.H<-pt %<+% div.df + geom_tippoint(aes(color=H),size=2) + theme_tree2() + scale_color_continuous(high='#D55E00',low='#0072B2')  +
    theme(legend.position="right") + ggtitle(paste0(treetitle," (Shannon evenness)")) 
  pt.H<-pt.H + annotate("text",x=1,y=5,label=paste0("Lambda: ",lambda.H))
 
 

    
  return(list(mean_divs,pt.asv,leg,pt.heat,pt.asvRF,pt.N,pt.H))
     }else{return(list(mean_divs))}
     }}

# [[1]]mean_divs, averaged alpha diversity scores
# [[2]]pt.asv, ASV tree plotted with abundances in hosts colored by host order
# [[3leg, legend
# [[4]]pt.heat, ASV abundance heatmap
# [[5]]pt.asvRF, ASVtree wtih hosts plotted as HMA/LMA 
# [[6]]pt.N, ASV tree number host habitats
# [[7]]pt.H, ASV tree Shannon H for habitats(host) mapped


#treefile<-trfiles[2]
plot_microbetree_hostrange<- function(treefile) {

  tr <- read.tree(treefile) 
  title<-unlist(strsplit(treefile,"\\."))[2]
  print(title)
  og<-grep("Outgroup",tr$tip.label)
  print(paste0("Outgroup: ",tr$tip.label[og]))
  rtr<-ladderize(root(tr,outgroup=og))
  #root then drop outgroups
  tr<-drop.tip(rtr,tip=og)
  anytips<-length(tr$tip.label[tr$tip.label %in% taxa_names(psa)])>2
  if(anytips==T){
    if(!file.exists(paste0("07_divresults/",title,".rds"))){
  res<-calc_div_signal(tree=tr,ps=psa, richness=S.ASV, shannons=shannons.ASV,treetitle=title)
  save(res,file=paste0(outDir,title,".rds"))
    }else{load(paste0(outDir,title,".rds"))}
  return(res)
  }
}

trfiles <- list.files(path="bact_trees/rootedtrees", pattern="RAxML_result*", full.names=TRUE, recursive=FALSE)

#set up table to capture summary...time consuming on large trees!

if(!file.exists("07_divresults/div_estimates.rds")){
    div_estimates<-lapply(trfiles,plot_microbetree_hostrange)
    xx<-unlist(lapply(trfiles,function(x) unlist(strsplit(x,"\\."))[2]))
    names(div_estimates)<-unlist(lapply( xx,function(x) strsplit(x,"_seqs")[[1]][1]))

    if(any(sapply(div_estimates, is.null)))
      {div_estimates<-div_estimates[-which(sapply(div_estimates, is.null))]}
  save(div_estimates,file=paste0(outDir,"div_estimates.rds"))

#pdf(paste0(outDir,"trees16S_diversities.pdf"))
#div_estimates
#dev.off()

# isolate plot types into separate files
pdf(paste0(outDir,"trees16S_xMA_occupancy.pdf"), height=11, width=8)
lapply(div_estimates,function (x) if( length(x)>1) x[[5]])
dev.off()
pdf(paste0(outDir,"trees16S_hostspecies.pdf"), height=11, width=8)
temp<-lapply(div_estimates,function (x) {if( length(x)>1) list(x[[2]],x[[3]])})
temp
dev.off()
pdf(paste0(outDir,"trees16S_heatmap.pdf"), height=11, width=8)
lapply(div_estimates,function (x) if( length(x)>1) x[[4]])
dev.off()
pdf(paste0(outDir,"trees16S_hostdiversitysignal.pdf"), height=11, width=8)
temp<-lapply(div_estimates,function (x) {if( length(x)>1) list(x[[6]],x[[7]])})
  temp
dev.off()

}else{load(paste0(outDir,"div_estimates.rds"))}
divind<-lapply (div_estimates, function(x) x[[1]]) #mean_divs N and H per bacterialgroup
bacterialgroup<-names(divind)

names(divind)<-NULL
out.df<-data.frame(matrix(unlist(divind),byrow=T,ncol=2,nrow=length(divind))); rownames(out.df)<-bacterialgroup
colnames(out.df)<-names(divind[[1]])

par(mfrow=c(2,2)) 
hist(out.df$N,breaks=20)
hist(out.df$H,breaks=20)

write.csv(out.df,file=paste0(outDir,"hostdiversity_indices.csv"))

```



##Abundance heatmaps: ASV trees vs sponge trees
```{r heats}
#plot heatmaps to show abundance across all phyla vs sponge tree
plot_heats<-function(trfile){
  tr <- read.tree(trfile) 
  title<-unlist(strsplit(trfile,"\\."))[2]
  title<-unlist(strsplit(title,"_seqs"))[1]
  
  print(title)
  # apply function
  og<-grep("Outgroup",tr$tip.label)
  
  print(paste0("Outgroup: ",tr$tip.label[og]))
  rtr<-ladderize(root(tr,outgroup=og))
  #root then drop outgroups
  #plot(rtrb)
  tr<-drop.tip(rtr,tip=og)
  anytips<-length(tr$tip.label[tr$tip.label %in% taxa_names(psa)])>2
  if(anytips==T){
  p.phylo<-ggtree::ggtree(tr,right=T) + ggtree::geom_tiplab()
  tipnames<-ggtree::get_taxa_name(p.phylo)
   if(any(sample_sums(psa)>0)){
    ps<-merge_phyloseq(psa,tr)
    ps<-prune_taxa(tr$tip.label,ps)
    sample_data(ps)$Species_consensus<-factor(sample_names(ps),levels=Species.order)
    #plot host occurence across ASV tips
    nonzerospp<-which(sample_sums(ps)>0)
    nonzerospp<- Species.order[Species.order %in% names(nonzerospp)]
    species_color<-treemeta.u$species_color[which(treemeta.u$Species_consensus %in% nonzerospp)]
    
    pt<-plot_tree(ps,size="abundance",color="Species_consensus",label.tips = "taxa_names") + scale_color_manual(values=c(species_color,"gray","black"))
    
    pt.asv<-pt + theme(legend.position="none") + ggtitle(title )
    legend <- cowplot::get_legend(pt ); leg<-cowplot::ggdraw(legend); 
    pt.heat<-plot_heatmap(ps,taxa.order=tipnames,sample.order=Species.order,title=paste0(title,": observed in sponges"))
    
    return(list(p.phylo,pt.asv,leg,pt.heat))
  }}
}


heats<-lapply(trfiles,plot_heats)
pdf(paste0(outDir,"heatmaps.pdf"))
heats
dev.off()

#using diversity measures saved in out.df, test if H, D, p, N are different between
# -HMA and LMA sponges.
# -different btween HMA and LMA type microbes (use top predicted from RF)


```



##Focus on Chloroflexi Dehalococcoidea heatmap
```{r chloroflex}
tr <- read.tree("bact_trees/rootedtrees/RAxML_result.Bacteria_Chloroflexi_Dehalococcoidia_seqs") 
  title<-"Chloroflexi_Dehalococcoidia"
  # apply function
  og<-grep("Outgroup",tr$tip.label)
  
  print(paste0("Outgroup: ",tr$tip.label[og]))
  rtr<-ladderize(root(tr,outgroup=og))
  tr<-drop.tip(rtr,tip=og)
  
  #identify focal subclades and prune off smaller trees
  plot(tr,tip.labels=F);nodelabels()
  library(phangorn)
  zoom(tr,focus=unlist(Descendants(tr,node=920,type="tips")))
par(mfrow=c(1,1))
trd1<-drop.tip(tr,tip=unlist(Descendants(tr,node=920,type="tips")))
tr1<-extract.clade(tr,node=920); length(tr1$tip.label); plot(tr1)
tr1b<-extract.clade(tr1,node=417); length(tr1b$tip.label); plot(tr1b)

plot(trd1); nodelabels(cex=0.5)
#505 and 676
tr2<-extract.clade(trd1,node=505); length(tr2$tip.label); plot(tr2)
tr3<-extract.clade(trd1,node=676); length(tr3$tip.label); plot(tr3)
tr3b<-extract.clade(tr3,node=317); length(tr3b$tip.label); plot(tr3b)

#need half as many tips
  plotsubtree<-function(subtr){
    p.phylo<-ggtree::ggtree(subtr,right=T) + ggtree::geom_tiplab(size=2)
  tipnames<-ggtree::get_taxa_name(p.phylo)
    ps<-merge_phyloseq(psa,subtr)
    ps<-prune_taxa(subtr$tip.label,ps)
    sample_data(ps)$Species_consensus<-factor(sample_names(ps),levels=Species.order)
    #plot host occurence across ASV tips
    nonzerospp<-which(sample_sums(ps)>0)
    nonzerospp<- Species.order[Species.order %in% names(nonzerospp)]
    species_color<-treemeta.u$species_color[which(treemeta.u$Species_consensus %in% nonzerospp)]
    
    pt<-plot_tree(ps,size="abundance",color="Species_consensus",label.tips = "taxa_names") + scale_color_manual(values=c(species_color,"gray","black"))
    
    pt.asv<-pt + theme(legend.position="none") + ggtitle(title )
    legend <- cowplot::get_legend(pt ); leg<-cowplot::ggdraw(legend); 
    pt.heat<-plot_heatmap(ps,taxa.order=rev(tipnames),sample.order=Species.order,title=paste0(title,": observed in sponges"))
  return(list(p.phylo,pt.asv,leg,pt.heat))
  }
  t1.plots<-plotsubtree(tr1b)
    t2.plots<-plotsubtree(tr2)
  t3.plots<-plotsubtree(tr3b)

  pdf(paste0(outDir,"ChloroflexiHeat.pdf"))
  t1.plots
  t2.plots
  t3.plots
  dev.off()
```



